---
title: "Data Analytics for Boston 311, year 2024"
author: "Joël R. Pradines, Ph.D."
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    math: true
    embed-resources: true
    link-external-newwindow: true
bibliography: references.bib
nocite: |
  @*
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(readr)
library(ggplot2)
library(dplyr)
library(lubridate)
library(ggplot2)
library(here)
library(tidyr)
```

```{r}
#| label: ingest
#| message: false
#| warning: false
#| echo: false

boston_311 <- read_csv(here("data", "tmpm461rr5o.csv"))
```

# 1. Introduction: Applied Analytics for Maintenance Operations

Managing large-scale property portfolios—spanning hundreds of buildings and thousands of service requests—requires more than basic descriptive dashboards. Reporting metrics like average completion times or SLA (Service Level Agreement) compliance rates frequently mask bottlenecks. They obscure capital constraints, disguise inefficient resource allocation, and fail to measure true workforce capacity.

Optimizing maintenance operations at scale requires advanced Data Science (DS) and Operations Research (OR) methodologies. Extracting actionable intelligence from service databases is an iterative process relying on three core ingredients:

**Statistical Rigor**: Averages are mathematically flawed in maintenance data due to survivorship bias and right-skewed extremes that prohibit normal assumptions. Robust probabilistic modeling is required to measure true performance and system stability, demanding deep statistical expertise.

**Human Detective Work**: Every operational dataset contains administrative artifacts, batch-processing illusions, and mislabeled text. Uncovering the truth requires an exploratory analytical process that generalized software cannot fully automate.

**Advanced Programming**: Moving beyond spreadsheets to deploy multivariate regressions, spatial autocorrelation, and localized machine learning pipelines is essential. While AI-aided programming accelerates this process, it still requires deep architectural understanding and practical implementation experience.

## The Example Dataset (Boston 311)

To demonstrate the application and value of these methodologies, this report utilizes the publicly available Boston 311 (Year 2024) dataset as an applied proxy.

Boston’s geographic districts and digital queues serve as a practical analog for decentralized property management. Just as a real estate portfolio is divided into building subsets managed by local teams, the city's workload is distributed across distinct neighborhoods. By treating constituent requests as tenant service orders, this document provides a demonstrative case study. While every new dataset is unique and requires its own custom detective work, this report illustrates how applying advanced analytics uncovers local bottlenecks, evaluates geographic equity, and models workforce dynamics.

## Key findings


# 2. Data Validation and Formatting

Data validation is the foundation of rigorous analysis. Before formatting operational features, we programmatically audit the raw dataset to map column types, quantify unique values, and identify missing records (NAs). This prevents silent failures in downstream mathematical models.

```{r}
#| label: tbl-data-dictionary
#| message: false
#| warning: false
#| echo: false
#| tbl-cap: "Dataset Overview: Column types, unique values, and missing data."

library(dplyr)
library(DT)

# 1. Overall dimensions
cat("Dataset Dimensions:", format(nrow(boston_311), big.mark = ","), "rows and", ncol(boston_311), "columns.\n\n")
# 2. Column-summary dataframe
data_summary <- tibble(
  Column = names(boston_311),
  Type = sapply(boston_311, function(x) paste(class(x), collapse = ", ")),
  Unique_Values = sapply(boston_311, n_distinct),
  Missing_Values = sapply(boston_311, function(x) sum(is.na(x)))
)
# 3. Interactive HTML table
datatable(
  data_summary,
  rownames = FALSE,
  options = list(
    pageLength = 10, 
    autoWidth = TRUE,
    dom = 'tp' # Keeps the interface clean (just table and pagination)
  ),
  caption = "Dataset Overview: Column types, unique values, and missing data."
)
```

Following the audit, we partition the raw data into two subsets to isolate different operational dynamics:

**Backlog Cohort (Open Requests)**: Work orders unresolved at the end of the observation period. This subset is critical for diagnosing structural blockages and system capacity.

**Resolved Cohort (Closed Requests)**: Work orders with a complete lifecycle. This primary dataset evaluates workforce speed and SLA compliance.

The pipeline below isolates the resolved cohort, purges data entry anomalies (e.g., negative resolution times), and generates new temporal features for subsequent seasonality and scheduling analysis.

```{r}
#| label: data-cleaning
#| message: false
#| warning: false
#| echo: false

# Ingestion
boston_311 <- read_csv(here("data", "tmpm461rr5o.csv"))
# Cleaning and Feature Engineering for the Resolved Cohort
boston_clean <- boston_311 %>%
  # Filter for completed work orders
  filter(case_status == "Closed", !is.na(closed_dt), !is.na(open_dt)) %>%
  mutate(
    # Exact resolution time
    resolution_days = as.numeric(difftime(closed_dt, open_dt, units = "days")),
    # Temporal features for seasonality and scheduling analysis
    open_month = month(open_dt, label = TRUE, abbr = TRUE),
    open_wday = wday(open_dt, label = TRUE, abbr = TRUE, week_start = 1), # Monday = 1
    open_hour = hour(open_dt)
  ) %>%
  # Remove data entry errors (negative resolution times)
  filter(resolution_days >= 0)
write_csv(boston_clean, here("data", "boston_clean.csv"))
# Sort output
cat("Cleaned dataset contains", format(nrow(boston_clean), big.mark = ","), "closed records ready for analysis.\n")
```

Before evaluating historical performance with the resolved cohort, we will first analyze the backlog cohort to diagnose the the system status at the end of 2024.

# 3. Open Requests at the End of 2024

We now dig into the subset of open requests. This will provide an example of how to stress-test a visual observation and detect a major operational disturbance.


```{r}
#| label: isolate-open-cases
#| message: false
#| warning: false
#| echo: false

# Tickets still open at the end of 2024
boston_open <- boston_311 %>%
  filter(case_status == "Open" | is.na(closed_dt)) %>%
  mutate(
    open_month = month(open_dt, label = TRUE, abbr = TRUE),
    # Ticket age as of the end of the 2024 calendar year
    days_open_at_yearend = as.numeric(difftime(as.POSIXct("2024-12-31 23:59:59"), open_dt, units = "days"))
  )
write_csv(boston_open, here("data", "boston_open.csv"))
# Short output
cat("Isolated", format(nrow(boston_open), big.mark = ","), "open records.\n\n")
```

@fig-open-tickets-distribution below overlays a density curve on top of a weekly-binned histogram. 

```{r}
#| label: fig-open-tickets-distribution
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Age distribution of open requests at the end of 2024."

ggplot(boston_open, aes(x = days_open_at_yearend)) +
  # Histogram with a 7-day binwidth to represent weekly operational cycles
  geom_histogram(aes(y = after_stat(density)), binwidth = 7, fill = "lightgray", color = "white") +
  # Density overlay to highlight the overall shape and search for bimodality
  geom_density(color = "darkred", linewidth = 1) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(0, 365, by = 60)) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Age Distribution of Open Requests at Year-End",
    subtitle = "Investigating bimodality: Active recent backlog vs. stagnant long-term requests.",
    x = "Days Open (as of Dec 31, 2024)",
    y = "Density (Proportion of Open Work Orders)"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )
```

Visual inspection suggests distinct peaks. A required step is questioning whether these represent real operational phenomena (like a seasonal surge) or are statistical artifacts caused by "binning bias" — the chosen width of the histogram buckets.

Because our sample size is large ($N > 40,000$), we can test the robustness of the observation by altering the histogram resolution. If the peaks are real, they will persist whether we group the data by day, week, or month.

```{r}
#| label: fig-robust-binning
#| fig-subcap: true
#| layout-ncol: 3
#| fig-width: 4
#| fig-height: 4
#| echo: false
#| warning: false
#| fig-cap: "Histograms of age distribution of open requests at the end of 2024, with different bin sizes: (a) days, (b) weeks and (c) months."

# Base plot for the first two charts mapping continuous days
p_base <- ggplot(boston_open, aes(x = days_open_at_yearend)) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(0, 365, by = 120)) +
  labs(x = "Days Open (as of Dec 31)", y = "Volume of Open Tickets") +
  theme(plot.title = element_text(size = 11, face = "bold"))
# Daily Binning
p_base + 
  geom_histogram(binwidth = 1, fill = "darkgray", color = "white", linewidth = 0.1) +
  labs(title = "Daily Resolution (1-Day Bins)")
# Weekly Binning
p_base + 
  geom_histogram(binwidth = 7, fill = "steelblue", color = "white", linewidth = 0.2) +
  labs(title = "Weekly Resolution (7-Day Bins)")
# Calendar Months
boston_open %>%
  # Reverse the month levels so December (newest) is on the left, matching the 0-365 days flow
  mutate(open_month_rev = factor(open_month, levels = rev(levels(open_month)))) %>%
  ggplot(aes(x = open_month_rev)) +
  geom_bar(fill = "darkred", color = "white", linewidth = 0.2) +
  theme_minimal() +
  labs(
    title = "Monthly Resolution (Calendar)",
    x = "Month Opened (Newest to Oldest)", 
    y = "Volume of Open Tickets"
  ) +
  theme(
    plot.title = element_text(size = 11, face = "bold"),
    # Angle the text slightly to ensure all 12 month labels fit cleanly
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

@fig-robust-binning suggests that the peaks at age 60-day and 170-day are robust. They do not seem to be binning artifacts. Yet, this statement is still based on visual inspection. So, the next step is to use a statistically rigorous test to check for peak presence.

Hartigan's Dip Test is a non-parametric test specifically designed to detect multimodality [@hartigan1985dip]. The null hypothesis ($H_0$) states that the distribution is unimodal. Mathematically, it calculates the "dip" statistic ($D$), which is the maximum vertical distance between the empirical distribution function, $F_n(x)$, and a modeled best-fitting unimodal distribution function, $U(x)$:
$$D = \sup_x \vert{}F_n(x) - U(x)\vert{}$$
If the data truly contains multiple structural waves (like the 60-day and 170-day spikes), $F_n(x)$ will deviate drastically from any possible unimodal curve and the $p$-value will approach zero, definitively rejecting the null hypothesis.


```{r}
#| label: hartigans-dip-test
#| echo: false
#| warning: false
#| message: false
#| results: hold

library(diptest)
library(dplyr)

# Test on the continuous days open variable
dip_result <- dip.test(boston_open$days_open_at_yearend)
# Extract and format the results
dip_stat <- round(dip_result$statistic, 4)
dip_pval <- format(dip_result$p.value, scientific = TRUE, digits = 3)
# Consolidate output into a single string
conclusion_text <- if(dip_result$p.value < 0.05) {
  "Conclusion: The p-value approaches zero, mathematically rejecting unimodality.\nThe distinct waves in the backlog are structural realities, not binning artifacts."
} else {
  "Conclusion: Failed to reject unimodality."
}

final_output <- paste0(
  "--- Hartigan's Dip Test for Unimodality ---\n",
  "Dip Statistic (D): ", dip_stat, "\n",
  "P-Value:           ", dip_pval, "\n\n",
  conclusion_text, "\n"
)
cat(final_output)
```


Having established the robustness of the 60-day and 170-day spikes, we investigate their potential origins by examining monthly requests that are still open at the end of 2024 by department in @fig-open-tickets-by-month.

```{r}
#| label: fig-open-tickets-by-month
#| echo: false
#| warning: false
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Visualization of departments contributions to open requests at the end of 2024 by starting month (Note that the month order is the reverse of that in @fig-robust-binning (c))."

# Identify the top 4 departments with the most open tickets
top_open_subjects <- boston_open %>%
  count(subject, sort = TRUE) %>%
  slice_head(n = 4) %>%
  pull(subject)
# Group the data for the stacked visualization
plot_data_open <- boston_open %>%
  mutate(
    subject_grouped = ifelse(subject %in% top_open_subjects, subject, "All Other Departments"),
    # Factor to ensure the grey baseline sits at the bottom of the stack
    subject_grouped = factor(subject_grouped, levels = c("All Other Departments", top_open_subjects))
  )
# Generate the bar chart
ggplot(plot_data_open, aes(x = open_month, fill = subject_grouped)) +
  geom_bar(color = "white", linewidth = 0.2) +
  scale_fill_manual(
    values = c(
      "All Other Departments" = "grey80",
      # Using a clear, distinct color palette for the top culprits
      setNames(c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3"), top_open_subjects)
    )
  ) +
  theme_minimal() +
  labs(
    title = "Year-End Backlog by Month of Origin",
    subtitle = "Exploratory view: Identifying seasonal waves and departmental bottlenecks.",
    x = "Month Ticket Was Opened",
    y = "Volume of Open Tickets at Year-End",
    fill = "Department"
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 9)
  ) +
  guides(fill = guide_legend(ncol = 2, reverse = TRUE))
```

The visualization explains the 170-day spike: the Transportation - Traffic Division backlog suddenly ballooned in June 2024 and remained saturated.

This blockage aligns perfectly with the **Sumner Tunnel Restoration Project**. MassDOT executed a full 24/7 closure from July 5 to August 5, 2024, with June dedicated to intense logistical preparation—altering traffic patterns, setting detours, and modifying transit routes. This monumental shift overwhelmed the Traffic Division's capacity, creating a localized backlog that persisted through year-end.

Beyond the June spike, the visualization reveals a secondary wave originating in October (the ~60-day backlog). Rather than a broad seasonal shift, this was a second, distinct shock striking the exact same department.

In October 2024, the MBTA executed a multi-week **shutdown of the Orange Line subway**, replacing trains with surface-level shuttle buses. To accommodate this, the Traffic Division rapidly implemented severe street-level modifications across the southwest corridor, including removing parking spaces, creating pop-up bus lanes, and altering traffic flow.

Like the summer closure, this logistical pivot overwhelmed the division. Routine neighborhood work orders were abandoned to manage the transit crisis, cementing a massive secondary backlog into the system right before the calendar year closed.

# 4. Resolution Time

Evaluating request resolution times using the average (mean) is a common pitfall. Maintenance completion times are rarely normally distributed. Instead, they exhibit a severe right skew: most tasks close quickly, while a handful of complex or disputed outliers remain open for months.

This long tail artificially inflates the mean, distorting the true operational baseline. Consequently, the median provides the mathematically sound metric for evaluating performance and establishing Service Level Agreements (SLAs).


```{r}
#| label: fig-resolution-density-log
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Density of Maintenance Resolution Times (Log Scale)"

library(scales) 

# Summary stats 
res_summary <- boston_clean %>%
  summarise(
    Mean = mean(resolution_days, na.rm = TRUE),
    Median = median(resolution_days, na.rm = TRUE)
  )
ggplot(boston_clean, aes(x = resolution_days)) +
  geom_density(fill = "steelblue", color = "darkblue", alpha = 0.5) +
  geom_vline(xintercept = res_summary$Mean, color = "darkred", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = res_summary$Median, color = "black", linetype = "solid", linewidth = 1) +
  # log1p transformation safely handles 0-day resolutions
  scale_x_continuous(
    trans = "log1p",
    breaks = c(0, 1, 3, 10, 30, 100, 365),
    labels = label_comma()
  ) +
  # Setting y = Inf and adjusting vjust pins the labels to the top of the plot
  annotate("text", x = res_summary$Mean, y = Inf, 
           label = paste("Mean:\n", round(res_summary$Mean, 1), "days"), 
           color = "darkred", hjust = -0.2, vjust = 1.5, size = 4, fontface = "bold") +
  annotate("text", x = res_summary$Median, y = Inf, 
           label = paste("Median:\n", round(res_summary$Median, 1), "days"), 
           color = "black", hjust = 1.2, vjust = 1.5, size = 4, fontface = "bold") +
  theme_minimal() +
  labs(
    title = "Density of Maintenance Resolution Times (Log Scale)",
    subtitle = "A log1p scale stretches the immediate closures while revealing the true span of the multi-month right tail.",
    x = "Days to Close (Log Scale)",
    y = "Density of Work Orders"
  )
```

As the density visualization illustrates, even after applying a logarithmic transformation, the distribution remains heavily right-skewed and multi-modal. The data fails to achieve log-normality. This mathematical reality precludes the use of standard parametric statistical models (which assume a normal distribution of errors) and necessitates the advanced, non-parametric, and survival-based methodologies deployed in the next sections.

# 5. SLA Compliance Analytics

Absolute resolution times only tell half the story. To truly evaluate vendor and operational performance, we must compare the completion timestamp (`closed_dt`) against the algorithmic deadline (`sla_target_dt`). 

This section isolates the subset of closed requests with defined service level agreements. We calculate a compliance rate (the percentage of tickets closed on or before their target date) grouped by the top service categories.

## 5.1 The SLA Delta Distribution (Lateness/Earliness)

We now analyze the distribution of the SLA Delta (Actual Resolution Time minus SLA Target Time). A value of 0 means the ticket was closed exactly on the deadline. Negative values indicate early closures, while positive values indicate the ticket was overdue.

```{r}
#| label: fig-sla-delta-distribution
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Distribution of SLA Delta."

# Calculate the exact delta in days
sla_delta_data <- boston_clean %>%
  filter(!is.na(sla_target_dt), subject != "Animal Control") %>%
  mutate(
    sla_delta_days = as.numeric(difftime(closed_dt, sla_target_dt, units = "days")),
    # Categorize for the plot fill
    compliance_status = ifelse(sla_delta_days <= 0, "On Time (Early)", "Overdue (Late)")
  )
# Plot the distribution of the delta
ggplot(sla_delta_data, aes(x = sla_delta_days, fill = compliance_status)) +
  # boundary = 0 added here to prevent overlapping bins at zero
  geom_histogram(binwidth = 1, boundary = 0, color = "white", alpha = 0.8) +
  # Add the hard SLA deadline reference line
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("On Time (Early)" = "steelblue", "Overdue (Late)" = "darkred")) +
  # Zoom the x-axis to a relevant operational window (e.g., 30 days early to 30 days late)
  coord_cartesian(xlim = c(-30, 30)) +
  theme_minimal() +
  labs(
    title = "Distribution of SLA Delta",
    subtitle = "The variance of actual closure times relative to their expected deadlines.",
    x = "SLA Delta (Days)",
    y = "Volume of Work Orders",
    fill = "Status"
  ) +
  theme(legend.position = "top")
```

Looking at this continuous distribution reveals the variance and behavioral clustering around the deadline. @fig-sla-delta-distribution shows a massive concentration of work orders closed immediately prior to the deadline, a sharp drop-off precisely at zero, and a long, right-skewed tail of overdue tickets.

## 5.2 Factors Driving Variance in SLA Compliance

To identify which factors drive early or late resolutions, we test the dependency between our categorical features (e.g., department, queue, geography,...) and the continuous SLA Delta. 

Because the dataset contains right-tail outliers, traditional variance-based tests (like an ANOVA) are not statistically sound to apply. In addition, the large $N$ value guarantees that all $p$-values approach zero, and the extreme skew breaks the underlying assumptions of ordinary least squares (OLS), rendering variance explained ($R^2$) unreliable.

### 5.2.1 The Mathematical Approach: Volume-Weighted K-S Distance

To correctly capture the true operational reality, we abandon variance and compare the **Empirical Cumulative Distribution Functions (ECDFs)** using a custom Volume-Weighted Kolmogorov-Smirnov (K-S) Statistic.

First, we define $F_{global}(x)$ as the ECDF of the SLA Delta for the entire city-wide portfolio. For any given factor (e.g., Department), we split the data into $K$ individual levels (subgroups). We then define $F_k(x)$ as the ECDF for a specific subgroup $k$.

The standard K-S statistic $D_k$ measures the maximum absolute vertical distance between the subgroup's ECDF and the global ECDF:

$$D_k = \sup_x |F_k(x) - F_{global}(x)|$$

This outputs a bounded value between $0$ (identical distributions) and $1$ (zero overlap). Because it relies purely on rank and percentiles, it is called a nonparametric statistic and it is immune to extreme outliers.

```{r}
#| label: fig-ks-statistic-explanation
#| echo: false
#| warning: false
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Graphical explanation of the Kolmogorov-Smirnov statistic."
# 1. Identify a "Goldilocks" subgroup (high volume, moderate D_k around 0.35)
global_dist <- sla_delta_data$sla_delta_days

illustrative_group <- sla_delta_data %>%
  group_by(type) %>%
  filter(n() > 5000) %>% 
  summarise(
    ks_stat = suppressWarnings(ks.test(sla_delta_days, global_dist)$statistic),
    .groups = "drop"
  ) %>%
  mutate(diff_from_target = abs(ks_stat - 0.35)) %>%
  arrange(diff_from_target) %>%
  slice(1)

max_type_name <- illustrative_group$type
max_type_dist <- sla_delta_data %>% filter(type == max_type_name) %>% pull(sla_delta_days)

# 2. Calculate the exact point of maximum vertical distance
x_vals <- sort(unique(c(global_dist, max_type_dist)))
ecdf_global_vals <- ecdf(global_dist)(x_vals)
ecdf_max_vals <- ecdf(max_type_dist)(x_vals)

diffs <- abs(ecdf_global_vals - ecdf_max_vals)
max_idx <- which.max(diffs)

x_max_diff <- x_vals[max_idx]
y_global <- ecdf_global_vals[max_idx]
y_max <- ecdf_max_vals[max_idx]
max_d_stat <- diffs[max_idx]

# Define dynamic labels to map math notation to the data
global_label <- "F_global : Global Baseline"
subgroup_label <- paste0("F_k : '", max_type_name, "' (Factor: type)")

# 3. Create the dataset for plotting
plot_data <- sla_delta_data %>%
  mutate(Curve = global_label) %>%
  bind_rows(
    sla_delta_data %>%
      filter(type == max_type_name) %>%
      mutate(Curve = subgroup_label)
  )

# 4. Generate the explanatory plot
ggplot(plot_data, aes(x = sla_delta_days, color = Curve)) +
  stat_ecdf(linewidth = 1) +
  # Add the vertical segment illustrating D_k
  annotate("segment", x = x_max_diff, xend = x_max_diff,
           y = y_global, yend = y_max,
           color = "darkred", linewidth = 1.2, 
           arrow = arrow(ends = "both", length = unit(0.1, "inches"))) +
  # Annotate the exact D_k value mathematically
  annotate("text", x = x_max_diff + 2, y = mean(c(y_global, y_max)),
           label = paste0("D[k] == ", round(max_d_stat, 3)), parse = TRUE,
           color = "darkred", hjust = 0, size = 5, fontface = "bold") +
  # Reference line for the SLA deadline (0 days)
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", alpha = 0.5) +
  # Map colors directly to the new dynamic labels
  scale_color_manual(values = setNames(c("darkgray", "steelblue"), c(global_label, subgroup_label))) +
  # Lock the zoom back to the critical operational window
  coord_cartesian(xlim = c(-30, 30)) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  labs(
    title = "Visualizing the Kolmogorov-Smirnov Statistic",
    subtitle = "The D_k statistic represents the maximum absolute vertical distance between the global distribution\nand a specific subgroup (in this example, the 'Parking Enforcement' level of the 'type' factor).",
    x = "SLA Delta (Days)",
    y = "Cumulative Proportion of Tickets Closed",
    color = NULL
  ) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 11, face = "bold")
  )
```

### 5.2.2 Results: The Dominance of Organizational Routing

To evaluate the *overall* explanatory power of the categorical factor, we must aggregate the $D_k$ scores of its levels. A simple average would mathematically equate a queue with $10$ tickets to a queue with $10,000$ tickets. Therefore, we calculate the **Volume-Weighted K-S Score** ($D_{weighted}$) by weighting each subgroup's K-S statistic by its proportional volume:

$$D_{weighted} = \sum_{k=1}^{K} \left( \frac{n_k}{N} \right) D_k$$

Where $n_k$ is the total number of work orders in subgroup $k$, and $N$ is the total volume across all subgroups. This penalizes noisy variance from tiny subgroups and rewards factors where massive subsets of work fundamentally deviate from the global baseline.


```{r}
#| label: feature-dependency-ks
#| echo: false
#| warning: false
#| message: false

library(purrr)
library(dplyr)

# Global baseline distribution
global_dist <- sla_delta_data$sla_delta_days
total_n <- nrow(sla_delta_data)
factors_to_test <- c("subject", "reason", "type", "queue", "department", 
                     "fire_district", "pwd_district", "city_council_district", 
                     "police_district", "neighborhood", "neighborhood_services_district", 
                     "ward", "precinct", "location_zipcode")
# Function to calculate the Weighted K-S Distance and extract p-values
calc_weighted_ks <- function(factor_name, data, global_dist, total_n) {
  data %>%
    group_split(!!sym(factor_name)) %>%
    map_dfr(function(df) {
      level_dist <- df$sla_delta_days
      level_n <- nrow(df)
      # Skip virtually empty levels to prevent test errors
      if(level_n < 5) return(tibble(level_n = level_n, ks_stat = 0, p_val = 1))
      # Suppress warnings for expected ties
      ks_res <- suppressWarnings(ks.test(level_dist, global_dist))
      tibble(
        level_n = level_n,
        ks_stat = ks_res$statistic,
        p_val = ks_res$p.value
      )
    }) %>%
    summarise(
      Factor = factor_name,
      Weighted_KS_Score = sum(ks_stat * (level_n / total_n)),
      Median_P_Value = median(p_val)
    )
}
# Map across all factors and rank
ks_results <- map_dfr(factors_to_test, ~calc_weighted_ks(.x, sla_delta_data, global_dist, total_n)) %>%
  arrange(desc(Weighted_KS_Score)) %>%
  mutate(
    Weighted_KS_Score = round(Weighted_KS_Score, 4),
    Median_P_Value = format(Median_P_Value, scientific = TRUE, digits = 3)
  )
```

Applying this non-parametric ranking to all organizational, task-based, and geographic features reveals a clear hierarchy of variance drivers displayed in @tbl-factor-ranking.


```{r}
#| label: tbl-factor-ranking
#| tbl-cap: "Factor Dependency Ranking (Weighted K-S Distance)"
#| echo: false
#| warning: false

library(knitr)

# Generate the formal Quarto table from the previous ks_results tibble
kable(
  ks_results, 
  col.names = c("Factor", "Weighted K-S Score", "Median P-Value"),
  align = c('l', 'c', 'c')
)
```


**Task & Routing Dominance**: The top drivers of distributional shift (queue, type, subject, and department) are entirely organizational. The physical nature of a request and its assigned digital routing queue fundamentally alter its resolution timeline compared to the global baseline.

**WARNING: The Geographic Equity Fallacy**: Spatial features (precinct, ward, location_zipcode, and municipal districts) rank uniformly at the bottom of the K-S results. A common **statistical mistake** is interpreting this low ranking as proof of the null hypothesis—falsely concluding that because geographic features do not explain the variance, geographic equity exists. The only correct statement for a large p-value is: "The test failed to reject the null hypothesis". Failure to reject is not equivalent to proving that the null hypothesis is true. Failing to reject the null hypothesis here only proves that slicing data by arbitrary administrative boundaries is a poor global predictor of SLA variance. Categorical bins destroy physical topology; a zipcode variable possesses no mathematical awareness of its neighbors. Accurately measuring geographic equity requires evaluating continuous spatial relationships via spatial autocorrelation, which we explicitly test in Section 10.

### 5.2.3 Result Illustration with Histograms

While the K-S statistic is mathematically robust against outliers, it is an abstract metric. To ground this finding back, we can look directly at the raw data for one of the highest-scoring factors: queue (D-score = 0.373).

By coloring the overall SLA Delta histogram by the top digital queues, we visually validate what the K-S test just measured. The distinct colors do not overlap uniformly; rather, they form separate, distinct operational distributions spanning from early completions to massive late backlogs.

```{r}
#| label: fig-queue-histogram
#| echo: false
#| warning: false
#| fig-width: 9
#| fig-height: 7
#| fig-cap: "SLA Delta Histogram"

# Isolate the top 4 queues to compare against the global background
top_queues <- sla_delta_data %>%
  count(queue, sort = TRUE) %>%
  slice_head(n = 4) %>%
  pull(queue)
# Group all remaining data into a global baseline category
plot_data_queue <- sla_delta_data %>%
  mutate(
    queue_grouped = ifelse(queue %in% top_queues, queue, "All Other Queues"),
    # Convert to factor and put "All Other Queues" first so it renders at the bottom of the stack
    queue_grouped = factor(queue_grouped, levels = c("All Other Queues", top_queues))
  )
# Generate the stacked histogram
ggplot(plot_data_queue, aes(x = sla_delta_days, fill = queue_grouped)) +
  geom_histogram(binwidth = 1, boundary = 0, position = "stack", color = "white", linewidth = 0.2, alpha = 0.9) +
  # Add the SLA deadline reference line
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 1) +
  # Zoom tightly to the central mass for higher resolution
  coord_cartesian(xlim = c(-15, 15)) +
  # Manually assign colors: neutral grey for the baseline, vibrant colors for the top 4
  scale_fill_manual(
    values = c(
      "All Other Queues" = "grey80",
      setNames(c("#E41A1C", "#377EB8", "#4DAF4A", "#FF7F00"), top_queues)
    )
  ) +
  theme_minimal() +
  labs(
    title = "SLA Delta Histogram by Top Queues",
    subtitle = "Visualizing the K-S finding: Top queues form distinct distributions against the global baseline.",
    x = "SLA Delta (Days)",
    y = "Volume of Work Orders",
    fill = "Digital Queue"
  ) +
  theme(
    # Move legend to the bottom to allow more horizontal breathing room
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 9)
  ) +
  # Force the legend to wrap into 2 columns so long department names are never cut off
  guides(fill = guide_legend(ncol = 2, reverse = TRUE))
```

# 6. Unsupervised Text Classification (BERTopic & Ollama)

To uncover the operational themes driving SLA delays without human bias or arbitrary keyword guessing, we deployed an unsupervised, locally hosted deep learning pipeline.

Using the sentence-transformers architecture [@reimers2019] on Apple Silicon, we converted unstructured closure notes into high-dimensional embeddings, clustering them via HDBSCAN [@mcinnes2017hdbscan].

Because raw statistical labels often lack semantic clarity, we integrated a local Large Language Model (Llama 3 via Ollama [@ollama2024]) to translate these clusters into clean, executive-ready categories.

Given the computational weight of Transformer models, the initial Python classification script was executed asynchronously. The results were saved to a static dataset, ensuring the compilation of this report remains lightweight.

### 6.1 The Python AI Pipeline

We utilized BERTopic [@grootendorst2022] combined with a [Hugging Face embedding model](https://huggingface.co/sentence-transformers/all-MiniLM-L6-v2) to group the closure notes into distinct operational bottlenecks.

### 6.2 Visualizing the AI Clusters

With the AI having categorized the unstructured text, we can now visualize the exact volume of delayed tickets attributed to each systemic issue. We apply a lightweight text-cleaning function in R to make the raw c-TF-IDF (Class-based Term Frequency-Inverse Document Frequency) labels readable. Unlike standard TF-IDF which evaluates individual documents, c-TF-IDF treats all documents in a single cluster as a single super-document, mathematically extracting the words that make that specific operational cluster unique compared to all others.

(Note: BERTopic assigns a topic ID of -1 to statistical outliers—tickets with highly unique text that do not fit into any broader operational pattern).

```{r}
#| label: fig-plot-bertopic-results
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Volume of SLA breaches categorized by AI-discovered operational themes."

library(readr)
library(dplyr)
library(ggplot2)
library(stringr)
library(here)

# Read ONLY the newly generated AI columns and the Ticket ID
ai_results <- read_csv(here("data", "boston_bertopic_results.csv")) %>%
  select(case_enquiry_id, topic_id, ai_topic_name) 
# Join the AI topics to your existing SLA dataset and filter
ai_plot_data <- sla_delta_data %>%
  inner_join(ai_results, by = "case_enquiry_id") %>%
  filter(compliance_status == "Overdue (Late)", topic_id != -1) %>%
  mutate(
    clean_topic = str_remove(ai_topic_name, "^\\d+_"),
    clean_topic = str_replace_all(clean_topic, "_", " "),
    clean_topic = str_to_title(clean_topic)
  ) %>%
  count(clean_topic, sort = TRUE) %>%
  slice_head(n = 15)
# Generate the visualization
ggplot(ai_plot_data, aes(x = n, y = reorder(clean_topic, n))) +
  geom_col(fill = "steelblue", color = "white", alpha = 0.9) +
  theme_minimal() +
  labs(
    title = "Top Causes of Operational Delays (Discovered via AI)",
    subtitle = "Unsupervised clustering of closure notes reveals distinct systemic bottlenecks.",
    x = "Volume of Delayed Work Orders",
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.y = element_text(size = 11, face = "bold")
  )
```

## 6.3 Refining Labels with Local Generative AI (Ollama)

While the plots above successfully utilize the mathematical c-TF-IDF keyword strings (e.g., "Related Responds Patrols Encourage"), these labels lack human context.

To bridge the gap between machine-readable statistics and executive-ready dashboards, we execute a 100% offline workflow using R's httr package to query a local Llama 3 instance via Ollama. This acts as an automated formatting bot, applying municipal context to translate the raw statistical outputs into polished categories. We will use these clean labels for subsequent analysis.

```{r}
#| label: tbl-llm-topic-translation
#| echo: false
#| warning: false
#| message: false
#| cache: true
#| tbl-cap: "Automated Topic Translation via Local Llama 3 (Offline API)"

library(dplyr)
library(purrr)
library(stringr)
library(readr)
library(here)
library(knitr)
library(httr)

# CLEAN FIRST, then get the Top 15 Unique topics
ai_results <- read_csv(here("data", "boston_bertopic_results.csv"))
topic_summary <- ai_results %>%
  filter(topic_id != -1) %>%
  mutate(
    clean_raw_topic = str_remove(ai_topic_name, "^\\d+_"),
    clean_raw_topic = str_replace_all(clean_raw_topic, "_", " ")
  ) %>%
  count(clean_raw_topic, sort = TRUE) %>%
  slice_head(n = 15)
# Direct Local API with the Proper Persona
translate_topic <- function(messy_string) {
  sys_prompt <- "You are an expert municipal data analyst for the City of Boston. You classify messy 311 work order keywords."
  user_prompt <- paste0(
    "Translate this messy keyword string into a clean, 3-to-4 word operational category label (e.g., 'Equipment Maintenance', 'Duplicate Request').\n\n",
    "Return ONLY the label. Do not include quotes, formatting, or introductory text.\n\n",
    "Keyword: ", messy_string
  )
  # Send offline request directly to local Mac memory
  res <- tryCatch({
    POST(
      url = "http://localhost:11434/api/generate",
      body = list(
        model = "llama3",
        prompt = user_prompt,
        system = sys_prompt,
        stream = FALSE
      ),
      encode = "json",
      timeout(30)
    )
  }, error = function(e) return(NULL))
  if (is.null(res) || status_code(res) != 200) return("API Connection Failed")
  raw_text <- content(res, as = "parsed", type = "application/json")$response
  # Strip out any markdown or quotes the LLM tries to sneak in
  clean_label <- str_remove_all(raw_text, '["*]') %>% str_trim()
  return(as.character(clean_label))
}
# Loop through the 15 topics
llm_clean_names <- map_chr(topic_summary$clean_raw_topic, translate_topic)
# Build and print the crosswalk table
topic_crosswalk <- tibble(
  `Original c-TF-IDF String` = topic_summary$clean_raw_topic,
  `LLM Dashboard Label` = llm_clean_names
)
kable(topic_crosswalk, caption = "Automated Topic Translation via Local Llama 3 (Offline API)")
```

## 6.4 Departmental Topic Enrichment

Now equipped with clean, AI-generated operational labels, we can mathematically test how these unsupervised clusters map to actual municipal departments.

Using Standardized Pearson Residuals from a Chi-Square test, the heatmap in @fig-topic-factor-enrichment highlights which operational bottlenecks are statistically over-represented in specific departmental workflows.

```{r}
#| label: fig-topic-factor-enrichment
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Enrichment heatmap showing which AI topics are statistically over-represented across municipal departments. Department abbreviations: BTDT (Transportation), INFO (Mayor's Hotline), ISD (Inspectional Services), PARK (Parks & Recreation), PROP (Property Management), PWDx (Public Works)."

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(here)

# Load the AI results and join the LLM translated labels
ai_df <- read_csv(here("data", "boston_bertopic_results.csv")) %>%
  filter(topic_id != -1, !is.na(department)) %>%
  mutate(
    clean_raw_topic = str_remove(ai_topic_name, "^\\d+_"),
    clean_raw_topic = str_replace_all(clean_raw_topic, "_", " ")
  ) %>%
  # Bring in the translations from the previous chunk
  left_join(topic_crosswalk, by = c("clean_raw_topic" = "Original c-TF-IDF String")) %>%
  mutate(
    # Fallback to the title-cased raw topic if it wasn't in our Top 15 translated list
    Final_Label = ifelse(is.na(`LLM Dashboard Label`), str_to_title(clean_raw_topic), `LLM Dashboard Label`)
  )
# Focus on top departments and top topics to keep the grid readable
top_depts  <- ai_df %>% count(department, sort = TRUE) %>% slice_head(n = 6) %>% pull(department)
top_topics <- ai_df %>% count(Final_Label, sort = TRUE) %>% slice_head(n = 12) %>% pull(Final_Label)
filtered_df <- ai_df %>%
  filter(department %in% top_depts, Final_Label %in% top_topics)
# Create a contingency table and run Chi-Square to get Standardized Residuals
contingency_matrix <- table(filtered_df$Final_Label, filtered_df$department)
chisq_res <- chisq.test(contingency_matrix)
# Convert residuals back to a tidy dataframe for ggplot
enrichment_data <- as.data.frame(chisq_res$stdres) %>%
  rename(Topic = Var1, Department = Var2, StdResidual = Freq)
# Visualize as an Enrichment Heatmap using the clean LLM labels
ggplot(enrichment_data, aes(x = Department, y = Topic, fill = StdResidual)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(
    low = "#2b83ba", mid = "#ffffbf", high = "#d7191c",
    midpoint = 0,
    name = "Standardized\nResidual"
  ) +
  theme_minimal() +
  labs(
    title = "Statistical Enrichment of AI Topics Across Departments",
    subtitle = "Red indicates strong over-representation (residual > +2); Blue indicates under-representation.",
    x = NULL,
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(size = 11, face = "bold")
  )
```

The results demonstrate clear organizational clustering: "Tree Pruning Operations" is intensely concentrated within the Parks Department (PARK), while "Cart Maintenance" heavily dominates Public Works (PWDX). Additionally, "Customer Feedback" shows a strong operational enrichment within the Mayor's Hotline (INFO). These highly localized, deep-red signals confirm that the unsupervised NLP pipeline successfully extracted distinct, department-specific operational themes directly from the unstructured closure notes.

# 7. Seasonal Demand and System Capacity

To optimize workforce schedules, we must understand seasonal demand fluctuations. High demand alone is not problematic if the workforce is appropriately scaled to handle it.

The true operational insight lies in comparing total inbound ticket volume (Demand) against the volume of missed SLAs (System Failure) month over month. This reveals the moments when operational capacity is overwhelmed by seasonal surges.

## 7.1 Citywide Macro Trends

We first look at the aggregate stress placed on Boston's municipal infrastructure across the entire year.

```{r}
#| label: fig-seasonal-trends-global
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Citywide monthly inbound ticket demand versus the volume of resulting SLA breaches."

library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(scales)

# Base dataset with extracted month
seasonal_base <- sla_delta_data %>%
  mutate(created_month = month(open_dt, label = TRUE, abbr = TRUE))
# Global Aggregation
global_summary <- seasonal_base %>%
  group_by(created_month) %>%
  summarise(
    Total_Demand = n(),
    SLA_Breaches = sum(compliance_status == "Overdue (Late)", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Total_Demand, SLA_Breaches),
    names_to = "Metric",
    values_to = "Volume"
  ) %>%
  mutate(
    Metric = factor(Metric, levels = c("Total_Demand", "SLA_Breaches"), 
                    labels = c("Total Inbound Tickets (Demand)", "Missed SLAs (System Failure)"))
  )
# Build the global seasonal overlay
ggplot(global_summary, aes(x = created_month, y = Volume, group = Metric, color = Metric)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3) +
  scale_y_continuous(labels = comma) +
  scale_color_manual(values = c("Total Inbound Tickets (Demand)" = "#b3cde3", 
                                "Missed SLAs (System Failure)" = "#e41a1c")) +
  theme_minimal() +
  labs(
    title = "Citywide Seasonal Operational Stress Test",
    subtitle = "Tracking when inbound municipal demand outpaces workforce resolution capacity.",
    x = "Month Request Opened",
    y = "Ticket Volume",
    color = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "top",
    axis.text.x = element_text(face = "bold")
  )
```

## 7.2 Departmental Micro Trends

While the citywide view provides the baseline, operational failures are rarely distributed evenly. By breaking the seasonality down by the highest-volume departments, we can see how different seasonal events (e.g., winter storms vs. summer park usage) might trigger capacity thresholds across distinct municipal departments.

```{r}
#| label: fig-seasonal-trends-department
#| echo: false
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Departmental breakdown of monthly demand versus SLA breaches. Department abbreviations: BTDT (Transportation), INFO (Mayor's Hotline), ISD (Inspectional Services), PARK (Parks & Recreation), PROP (Property Management), PWDx (Public Works)."

# Find the top 6 departments by overall ticket volume 
top_depts_season <- seasonal_base %>%
  count(department, sort = TRUE) %>%
  slice_head(n = 6) %>%
  pull(department)
# Departmental Aggregation
dept_summary <- seasonal_base %>%
  filter(department %in% top_depts_season) %>%
  group_by(department, created_month) %>%
  summarise(
    Total_Demand = n(),
    SLA_Breaches = sum(compliance_status == "Overdue (Late)", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Total_Demand, SLA_Breaches),
    names_to = "Metric",
    values_to = "Volume"
  ) %>%
  mutate(
    Metric = factor(Metric, levels = c("Total_Demand", "SLA_Breaches"), 
                    labels = c("Total Inbound Tickets (Demand)", "Missed SLAs (System Failure)"))
  )
# Build the faceted visualization
ggplot(dept_summary, aes(x = created_month, y = Volume, group = Metric, color = Metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  facet_wrap(~ department, scales = "free_y", ncol = 3) +
  scale_y_continuous(labels = comma) +
  scale_color_manual(values = c("Total Inbound Tickets (Demand)" = "#b3cde3", 
                                "Missed SLAs (System Failure)" = "#e41a1c")) +
  theme_minimal() +
  labs(
    title = "Seasonal Stress Test by Top Departments",
    subtitle = "Comparing capacity thresholds across individual workflows (Note: Y-axes vary by department).",
    x = "Month Request Opened",
    y = "Ticket Volume",
    color = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11),
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## 7.3 The Limits of Unsupervised NLP on Seasonal Surges

Re-applying the text classification pipeline from Section 6 to diagnose these seasonal trends failed to produce statistically valid results.

While the unsupervised model successfully identified global operational bottlenecks, slicing the unstructured text into department-and-month-specific windows severely degraded the signal-to-noise ratio. At this granular level, the AI clusters were overwhelmed by generic administrative shorthand rather than distinct physical phenomena.

Because unstructured text mining could not reliably explain these seasonal surges, we shift from descriptive categorization to measuring the system's mathematical dynamics: analyzing workforce velocity and system survival rates in the following sections.
  
# 8. Operational Reliability (Survival Analysis)

While average resolution times are commonly used in municipal reporting, they suffer from a severe statistical flaw: survivorship bias. By excluding tickets that are currently open, standard averages artificially deflate the apparent resolution time of heavily backlogged departments.

## 8.1 Kaplan-Meier

To capture the true operational reliability of the city, we apply time-to-event Survival Analysis. Using the Kaplan-Meier estimator, we plot the exact probability that a constituent's request will remain unresolved as days pass, handling open tickets as right-censored data points.

```{r}
#| label: fig-survival-analysis-departments
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Kaplan-Meier survival curves estimating the probability a ticket remains open over time. Department abbreviations: BTDT (Transportation), INFO (Mayor's Hotline), ISD (Inspectional Services), PARK (Parks & Recreation), PROP (Property Management), PWDx (Public Works)."

library(dplyr)
library(ggplot2)
library(survival)
library(broom)
library(stringr)
library(scales)

# Isolate the top 5 departments
top_5_depts <- sla_delta_data %>%
  count(department, sort = TRUE) %>%
  slice_head(n = 5) %>%
  pull(department)
survival_df <- sla_delta_data %>%
  filter(department %in% top_5_depts) %>%
  mutate(
    max_date = max(open_dt, na.rm = TRUE),
    calc_end_dt = if_else(is.na(closed_dt), max_date, closed_dt),
    duration_days = as.numeric(difftime(calc_end_dt, open_dt, units = "days")),
    status_ind = if_else(is.na(closed_dt), 0, 1)
  ) %>%
  filter(duration_days >= 0)
# Fit the Kaplan-Meier Model
km_fit <- survfit(Surv(duration_days, status_ind) ~ department, data = survival_df)
tidy_km <- tidy(km_fit) %>% mutate(clean_dept = str_remove(strata, "department="))
# Visualize the Hazard Curves
ggplot(tidy_km, aes(x = time, y = estimate, color = clean_dept)) +
  geom_step(linewidth = 1.2, alpha = 0.8) +
  coord_cartesian(xlim = c(0, 60)) + 
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  labs(
    title = "Time-to-Resolution: Ticket Survival Analysis",
    subtitle = "The mathematical probability that a 311 request remains unresolved as days pass.",
    x = "Days Since Request Opened",
    y = "Probability Ticket is Still Open",
    color = "Department"
  ) +
  theme(plot.title = element_text(face = "bold", size = 14))
```

The Kaplan-Meier estimation reveals stark differences in how municipal departments process their queues, highlighting both highly reactive workforces and structural bottlenecks:

**Public Works (PWDx)**: The orange curve plummets almost instantly, indicating that roughly 80% to 85% of their tickets are resolved within the first 48 to 72 hours. This mathematically depicts a workforce optimized for rapid, high-volume triage (e.g., immediate pothole patching, trash pickup).

**Baseline Cluster (BTDT, INFO, ISD)**: Transportation, the Mayor's Hotline, and Inspectional Services track closely together. They successfully clear about 75% of their open tickets within the first week. After that initial drop, they demonstrate a steady, predictable burn-down rate for more complex cases, approaching a near-zero unresolved rate by the 60-day mark.

**Parks and Recreation (PARK)**: The purple curve is the critical anomaly. While PARK resolves about half of its tickets in the first few days, the curve abruptly flattens into a concrete floor. By day 20, nearly 30% of their tickets remain open, and approximately 20% are essentially abandoned past the 60-day mark. This mathematical "long-tail" confirms that PARK does not suffer from a simple speed issue, but rather a structural inability to clear complex tickets, heavily implying capital budget constraints or a reliance on seasonal workforce unlocks.

## 8.2 Multivariate Drivers of Resolution (Cox Proportional Hazards)

While the Kaplan-Meier curves show us the overall survival rate of tickets by department, municipal operations do not exist in a vacuum. A ticket's resolution time is simultaneously influenced by the department assigned, the method the constituent used to report it, and whether it includes rich context like photographs.To untangle these overlapping variables, we apply a Cox Proportional Hazards Model:
$$h(t\vert{}X) = h_0(t) \exp(\beta_1 X_1 + \beta_2 X_2 + \dots + \beta_p X_p)$$
Where:

* $t$ represents the time elapsed (in days) since the request was opened.
* $h(t\vert{}X)$ is the hazard rate. In this operational context, it is the probability of the ticket successfully closing at time $t$, given a specific set of operational variables $X$.
* $h_0(t)$ is the baseline hazard, representing the underlying resolution speed of the system when all variables are held at their default reference levels.
* $X_1, \dots, X_p$ are the categorical and continuous predictor variables (our covariates, such as department and reporting source).$\beta_1, \dots, \beta_p$ are the coefficients estimated by the regression model.

This multivariate regression exponentiates these coefficients ($\exp(\beta)$) to calculate a Hazard Ratio (HR) for each variable. In this context, a "hazard" is a positive event (closing the ticket).

* HR > 1: The variable mathematically accelerates resolution (e.g., an HR of 1.5 means the ticket is closed 50% faster than the baseline).
* HR < 1: The variable acts as a bottleneck, delaying resolution.

```{r}
#| label: fig-cox-regression-model-adjusted
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Hazard Ratios (Log Scale) isolating the impact of reporting sources and departments. Automated API integrations (e.g., 'Maximo Integration' used by the Boston 311 system) have been removed to isolate human workforce efficiency."

library(dplyr)
library(survival)
library(broom)
library(ggplot2)
library(forcats)
library(stringr)

# Prepare Data and Filter Automated Artifacts
cox_df <- sla_delta_data %>%
  filter(department %in% c("BTDT", "ISD", "INFO", "PARK", "PWDx")) %>%
  mutate(
    calc_end_dt = if_else(is.na(closed_dt), max(open_dt, na.rm = TRUE), closed_dt),
    duration_days = as.numeric(difftime(calc_end_dt, open_dt, units = "days")),
    status_ind = if_else(is.na(closed_dt), 0, 1),
    source_clean = if_else(is.na(source), "Unknown", as.character(source))
  ) %>%
  # Remove the Maximo API artifact to measure actual human work
  filter(duration_days > 0, !str_detect(source_clean, "Maximo"))
# Fit the Adjusted Cox Model
cox_model <- coxph(Surv(duration_days, status_ind) ~ department + source_clean, data = cox_df)
# Tidy and Exponentiate
tidy_cox <- tidy(cox_model, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "") %>%
  mutate(
    clean_term = str_remove_all(term, "department|source_clean"),
    is_significant = if_else(p.value < 0.05, "Significant (p < .05)", "Not Significant"),
    clean_term = fct_reorder(clean_term, estimate)
  )
# Build the Log-Scaled Forest Plot
ggplot(tidy_cox, aes(x = estimate, y = clean_term, color = is_significant)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black", linewidth = 1) +
  geom_point(size = 4) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, linewidth = 1.2) +
  # Apply base-10 log scale for mathematically accurate visual distances
  scale_x_log10(breaks = c(0.25, 0.5, 0.75, 1, 1.5, 2, 3, 5)) + 
  scale_color_manual(values = c("Significant (p < .05)" = "#1b9e77", "Not Significant" = "#7570b3")) +
  theme_minimal() +
  labs(
    title = "Multivariate Drivers of Ticket Resolution",
    subtitle = "Variables to the right (HR > 1) accelerate resolution. Variables to the left (HR < 1) delay it.",
    x = "Hazard Ratio (Log-Scaled Multiplier of Resolution Speed)",
    y = "Operational Variable",
    color = "Statistical Significance"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom",
    axis.text.y = element_text(face = "bold", size = 11)
  )
```

In the multivariate forest plot, the central dashed line at 1.00 establishes baseline resolution speed. Dots represent the Hazard Ratio (HR)—a direct speed multiplier. Variables to the right (HR > 1) accelerate resolution; those to the left (HR < 1) act as bottlenecks. The horizontal bars indicate 95% confidence intervals. Because none cross the baseline, we can state with 95% statistical confidence that these variables genuinely impact workforce velocity, rejecting the possibility of random anomalies.

Filtering out automated "Maximo Integration" API calls isolates true human labor, revealing that resolution speed heavily depends on departmental capacity and intake friction. 

* **Public Works (PWDx)** operates at nearly double the baseline speed, validating its rapid triage design
* **Parks (PARK)** faces a 35% speed deficit from long-tail bottlenecks. 
* **INFO**: manual intake channels like phone calls introduce administrative friction, slowing resolution pipelines by roughly 22% compared to digitally precise reporting.

# 9. System Stability: Queueing Theory and Backlog Velocity

While Survival Analysis evaluates the lifespan of individual tickets, it does not measure the aggregate dynamic health of a department. To diagnose systemic instability, we must shift from the ticket level to the macro system level using Queueing Theory. This approach relies strictly on the mathematical balance between the Arrival Rate ($\lambda$)—tickets opened—and the Service Rate ($\mu$)—tickets closed. By calculating the net weekly queue change, we can strip away subjective text and measure pure operational dynamics. When $\lambda > \mu$, the system enters a state of structural instability (growing backlog), regardless of how fast routine tickets are closed.

```{r}
#| label: fig-queue-velocity-analysis
#| echo: false
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Queueing Stability: Weekly net change in unresolved tickets (Arrivals minus Services).Department abbreviations: BTDT (Transportation), INFO (Mayor's Hotline), ISD (Inspectional Services), PARK (Parks & Recreation), PROP (Property Management), PWDx (Public Works)."

library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)

# Calculate the Arrival Rate (lambda) and Service Rate (mu) per week
queue_dynamics <- sla_delta_data %>%
  filter(department %in% c("BTDT", "ISD", "INFO", "PARK", "PWDx")) %>%
  mutate(
    open_week = floor_date(open_dt, "week"),
    close_week = floor_date(closed_dt, "week")
  )
# Arrivals per week
arrivals <- queue_dynamics %>%
  group_by(department, week = open_week) %>%
  summarise(tickets_opened = n(), .groups = "drop")
# Services per week (ignoring NAs for open tickets)
services <- queue_dynamics %>%
  filter(!is.na(close_week)) %>%
  group_by(department, week = close_week) %>%
  summarise(tickets_closed = n(), .groups = "drop")
# Merge and calculate Net Backlog Velocity
backlog_velocity <- arrivals %>%
  full_join(services, by = c("department", "week")) %>%
  # Replace NAs with 0 for weeks where no tickets were opened or closed
  replace_na(list(tickets_opened = 0, tickets_closed = 0)) %>%
  mutate(
    net_queue_change = tickets_opened - tickets_closed,
    # Create a flag for color coding: Red = Growing queue, Blue = Shrinking queue
    status = if_else(net_queue_change > 0, "Queue Growing (Instability)", "Queue Recovering (Stability)")
  ) %>%
  # Filter out extreme edge dates if necessary to keep the plot clean
  filter(week >= min(arrivals$week, na.rm = TRUE) & week <= max(arrivals$week, na.rm = TRUE))
# Visualize the System Stability
ggplot(backlog_velocity, aes(x = week, y = net_queue_change, fill = status)) +
  geom_col(alpha = 0.85) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  facet_wrap(~ department, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("Queue Growing (Instability)" = "#e41a1c", 
                               "Queue Recovering (Stability)" = "#377eb8")) +
  theme_minimal() +
  labs(
    title = "Departmental Queue Stability (Backlog Velocity)",
    subtitle = "Tracking weekly thermodynamic balance: Tickets Opened vs. Tickets Closed.",
    x = "Week",
    y = "Net Change in Open Tickets",
    fill = "System Status"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

The Backlog Velocity chart maps the behavioral pulse of the city's workforce without relying on subjective administrative text. It reveals distinct operational realities across the monitored departments:

**PARK**: For the first six months of the year, the Parks Department operates in an unbroken state of structural instability. The continuous red growth confirms they lack the baseline daily capacity to match inbound requests. The massive blue correction in July aligns perfectly with the July 1st start of the Boston municipal Fiscal Year. This strongly suggests a seasonal budget unlock, potentially allowing the department to finally deploy the capital or short-term seasonal workforce required to crush the six-month backlog.

**BTDT**: Transportation displays a massive, singular correction of over 1,500 closed tickets in late summer. Because human workforces do not organically resolve 1,500 complex street-level issues in a single week, this mathematical signature strongly suggests an artifact of database batch-closure. It might represent a forced administrative reset to clear a runaway queue rather than natural operational velocity.

**PWDx & INFO**: Public Works and the Mayor's Hotline constantly oscillate between small delays and immediate recoveries, closely hugging the baseline. This suggests highly reactive, well-managed workforces operating with zero structural slack. They seem to absorb immediate shocks (such as the January weather surge for PWDx) and immediately deploy surge capacity the following week to reset to zero.

**ISD**: Inspectional Services demonstrates a steady accumulation of backlog through the spring months, directly aligning with the seasonal surge in construction and housing permits. Sustained recovery does not begin until mid-summer, indicating a persistent, multi-month lag between peak citizen demand and peak service delivery.

# 10. Spatial Equity (Geospatial Autocorrelation)

While overall departmental velocity provides a measure of system health, it does not address the distribution of that service. A critical mandate of municipal operations is equitable service delivery across all neighborhoods.

To mathematically test if operational failures are equitably distributed or geographically biased, we apply spatial autocorrelation using Global Moran’s I [@moran1950notes; @li2007beyond]. By mapping the SLA breach rates (the percentage of tickets that miss their target resolution time) across Boston's spatial boundaries, we can identify statistically significant "cold spots"—clusters of neighborhoods that systematically receive delayed service compared to the citywide baseline.

## 10.1 Mapping SLA Breach Clusters

To mathematically test if operational failures are equitably distributed or geographically biased, we apply spatial autocorrelation using Global Moran’s I. The statistic compares the variance of SLA breach rates across neighboring areas against the overall citywide variance:
$$I = \frac{N}{W} \frac{\sum_i \sum_j w_{ij}(x_i - \bar{x})(x_j - \bar{x})}{\sum_i (x_i - \bar{x})^2}$$
Where:

* $N$ is the total number of spatial units (in this case, census tracts) indexed by $i$ and $j$.
* $x_i$ and $x_j$ are the SLA breach rates for tracts $i$ and $j$, respectively.
* $\bar{x}$ is the citywide average SLA breach rate.
* $w_{ij}$ is the spatial weight matrix determining how tract $i$ and tract $j$ are connected (typically $1$ if they share a physical border, and $0$ if they do not).
* $W$ is the sum of all spatial weights ($W = \sum_i \sum_j w_{ij}$).

An index near $0$ implies that service delays are completely random ("equitable failure"). A positive index approaching $1$ proves that SLA breaches are structurally clustered, creating "cold spots" of systematic neglect. To ensure we are measuring true human equity rather than software architecture, automated closures originating from the "maximo integration" API have been strictly filtered out of this spatial model.

```{r}
#| label: fig-spatial-equity-morans-i
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Geospatial distribution of SLA breach rates across Boston census tracts."

library(dplyr)
library(sf)
library(tigris)
library(spdep)
library(ggplot2)
library(stringr)

# Gets the Shapefile via API (Natively downloads Suffolk County Census Tracts)
options(tigris_use_cache = TRUE)
boston_tracts <- tracts(state = "MA", county = "Suffolk", cb = TRUE, year = 2022, progress_bar = FALSE) %>%
  st_transform(crs = 4326)
# Prepare the 311 Spatial Points
spatial_311 <- sla_delta_data %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  filter(!str_detect(if_else(is.na(source), "Unknown", as.character(source)), "Maximo")) %>%
  mutate(is_breach = if_else(on_time == "ONTIME", 0, 1)) %>% 
  # Convert flat data frame into an active spatial object
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
# Spatial Join: Assign tickets to Census Tracts and calculate breach rates
tract_breaches <- boston_tracts %>%
  st_join(spatial_311, left = FALSE) %>%
  group_by(GEOID) %>%
  summarise(
    total_tickets = n(),
    breach_rate = mean(is_breach, na.rm = TRUE)
  ) %>%
  # Filter out tracts with highly volatile low sample sizes (e.g., airport, harbor islands)
  filter(total_tickets > 50)
# Calculate Global Moran's I for Spatial Autocorrelation
# Create the spatial weights matrix (defining which tracts touch each other)
neighbors <- poly2nb(tract_breaches, queen = TRUE)
weights <- nb2listw(neighbors, style = "W", zero.policy = TRUE)
# Execute the mathematical test
moran_result <- moran.test(tract_breaches$breach_rate, weights, zero.policy = TRUE)
# Visualization
ggplot(tract_breaches) +
  geom_sf(aes(fill = breach_rate), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "magma", labels = scales::percent_format()) +
  theme_void() +
  labs(
    title = "Spatial Equity of Municipal Services",
    subtitle = paste0("Global Moran's I: ", round(moran_result$estimate[1], 3), 
                      " (p-value: ", signif(moran_result$p.value, 3), ")"),
    fill = "SLA Breach Rate",
    caption = "Excludes automated Maximo Integration closures to isolate human workforce equity."
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "right"
  )
```

@fig-spatial-equity-morans-i shows that the spatial distribution of service delays is far from random. The calculated Global Moran's I statistic of 0.25, coupled with a $p$-value of $1.37 \times 10^{-9}$, mathematically rejects the null hypothesis of spatial randomness. 

Visually, the map strongly corroborates this test: census tracts experiencing severe SLA breach rates (the bright orange and yellow zones exceeding 30%) are geographically contiguous, forming structural "hotspots" of operational delay. Conversely, neighborhoods receiving highly efficient service (the dark purple and black zones hovering around a 10% breach rate) form their own distinct, localized clusters. This definitive spatial autocorrelation proves that the promptness of municipal service is systematically tied to geographic location.

## 10.2 Localized Clusters of Neglect (LISA)

Having proven that municipal delays are not uniformly distributed, we apply Local Indicators of Spatial Association (LISA) [@anselin1995local] to mathematically draw boundaries around the city's operational "hotspots" and "coldspots." By scaling the breach rates and calculating their spatial lags, we can isolate census tracts that suffer from statistically significant, systemic neglect.

Crucially, because automated closures originating from the "Maximo Integration" API have been removed from this spatial model, the resulting clusters represent genuine disparities in human workforce deployment and street-level service delivery.

```{r}
#| label: fig-spatial-equity-lisa
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "LISA Cluster Map isolating statistically significant zones of systemic neglect (High-High) and exceptional service (Low-Low)."
# Calculate Local Moran's I
local_m <- localmoran(tract_breaches$breach_rate, weights, zero.policy = TRUE)
# Scale the variables to determine the quadrants (centered on the mean)
c_breach <- tract_breaches$breach_rate - mean(tract_breaches$breach_rate, na.rm = TRUE)
c_lag <- lag.listw(weights, c_breach, zero.policy = TRUE)
# Categorize the tracts into LISA clusters
lisa_tracts <- tract_breaches %>%
  mutate(
    # Extract the p-value from the local moran matrix (Column 5)
    p_val = local_m[, 5], 
    # Define the raw spatial quadrants
    quadrant = case_when(
      c_breach > 0 & c_lag > 0 ~ "High-High (Systemic Neglect)",
      c_breach < 0 & c_lag < 0 ~ "Low-Low (Exceptional Service)",
      c_breach > 0 & c_lag < 0 ~ "High-Low (Spatial Outlier)",
      c_breach < 0 & c_lag > 0 ~ "Low-High (Spatial Outlier)",
      TRUE ~ "Not Significant"
    ),
    # Mask the quadrants by statistical significance (p < 0.05)
    lisa_cluster = if_else(p_val <= 0.05, quadrant, "Not Significant")
  ) %>%
  # Convert the cluster to a factor for ordered plotting
  mutate(lisa_cluster = factor(lisa_cluster, levels = c(
    "High-High (Systemic Neglect)", 
    "Low-Low (Exceptional Service)", 
    "High-Low (Spatial Outlier)", 
    "Low-High (Spatial Outlier)", 
    "Not Significant"
  )))
# Map the LISA Clusters
# Define a stark, color-blind friendly palette to highlight the inequity
lisa_colors <- c(
  "High-High (Systemic Neglect)" = "#d73027",    # Red
  "Low-Low (Exceptional Service)" = "#4575b4",   # Blue
  "High-Low (Spatial Outlier)" = "#fdae61",      # Light Orange
  "Low-High (Spatial Outlier)" = "#abd9e9",      # Light Blue
  "Not Significant" = "#f0f0f0"                  # Light Grey
)
ggplot(lisa_tracts) +
  geom_sf(aes(fill = lisa_cluster), color = "darkgrey", size = 0.1) +
  scale_fill_manual(values = lisa_colors, drop = FALSE) +
  theme_void() +
  labs(
    title = "Zones of Systemic Neglect vs. Exceptional Service",
    subtitle = "Statistically significant spatial clusters of 311 SLA breaches (Local Moran's I, p < 0.05).",
    fill = "LISA Classification",
    caption = "Excludes automated 'maximo integration' closures to isolate human workforce equity."
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(face = "bold")
  )
```

@fig-spatial-equity-lisa reveals the precise neighborhood boundaries of systemic operational failures, moving beyond simple density to mathematically isolate zones of structural inequality.

The LISA analysis identifies three distinct "High-High" clusters of systemic neglect (highlighted in red) that are geographically separated from the city's central core:

**Western Cluster (Allston-Brighton)**: The largest contiguous zone of delayed service dominates the far western edge of the city.

**Northeast Cluster (East Boston)**: A secondary significant zone of systemic delay encompasses the northeast peninsula, bounded by Boston Harbor.

**Northern Cluster (Charlestown)**: A third, smaller pocket of concentrated SLA breaches is situated on the northern peninsula across the Charles River.

Conversely, the tracts classified as "Low-Low" (highlighted in blue)—indicating localized clusters of exceptionally prompt service delivery—are positioned centrally, situated within and immediately adjacent to the downtown municipal core. This spatial reality demonstrates that while service delays are not randomly distributed, they form distinct, localized geographic pockets that merit targeted operational review.

# Appendix: R and Python Code

## Section 1

```{r}
#| label: ingest-A
#| message: false
#| warning: false
#| echo: true
#| eval: false

boston_311 <- read_csv(here("data", "tmpm461rr5o.csv"))
```

## Section 2

```{r}
#| label: data-dictionary-A
#| message: false
#| warning: false
#| echo: true
#| eval: false

library(dplyr)
library(DT)

# Overall dimensions
cat("Dataset Dimensions:", format(nrow(boston_311), big.mark = ","), "rows and", ncol(boston_311), "columns.\n\n")
# Column-summary dataframe
data_summary <- tibble(
  Column = names(boston_311),
  Type = sapply(boston_311, function(x) paste(class(x), collapse = ", ")),
  Unique_Values = sapply(boston_311, n_distinct),
  Missing_Values = sapply(boston_311, function(x) sum(is.na(x)))
)
# Interactive HTML table
datatable(
  data_summary,
  rownames = FALSE,
  options = list(
    pageLength = 10, 
    autoWidth = TRUE,
    dom = 'tp' # Keeps the interface clean (just table and pagination)
  ),
  caption = "Dataset Overview: Column types, unique values, and missing data."
)
```

```{r}
#| label: data-cleaning-A
#| message: false
#| warning: false
#| echo: true
#| eval: false

# Ingestion
boston_311 <- read_csv(here("data", "tmpm461rr5o.csv"))
# Cleaning and Feature Engineering for the Resolved Cohort
boston_clean <- boston_311 %>%
  # Filter for completed work orders
  filter(case_status == "Closed", !is.na(closed_dt), !is.na(open_dt)) %>%
  mutate(
    # Exact resolution time
    resolution_days = as.numeric(difftime(closed_dt, open_dt, units = "days")),
    # Temporal features for seasonality and scheduling analysis
    open_month = month(open_dt, label = TRUE, abbr = TRUE),
    open_wday = wday(open_dt, label = TRUE, abbr = TRUE, week_start = 1), # Monday = 1
    open_hour = hour(open_dt)
  ) %>%
  # Remove data entry errors (negative resolution times)
  filter(resolution_days >= 0)
write_csv(boston_clean, here("data", "boston_clean.csv"))
# Sort output
cat("Cleaned dataset contains", format(nrow(boston_clean), big.mark = ","), "closed records ready for analysis.\n")
```

## Section 3

```{r}
#| label: isolate-open-cases-A
#| message: false
#| warning: false
#| echo: true
#| eval: false

# Tickets still open at the end of 2024
boston_open <- boston_311 %>%
  filter(case_status == "Open" | is.na(closed_dt)) %>%
  mutate(
    open_month = month(open_dt, label = TRUE, abbr = TRUE),
    # Ticket age as of the end of the 2024 calendar year
    days_open_at_yearend = as.numeric(difftime(as.POSIXct("2024-12-31 23:59:59"), open_dt, units = "days"))
  )
write_csv(boston_open, here("data", "boston_open.csv"))
# Short output
cat("Isolated", format(nrow(boston_open), big.mark = ","), "open records.\n\n")
```

```{r}
#| label: fig-open-tickets-distribution-A
#| echo: true
#| warning: false
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Age distribution of open requests at the end of 2024."
#| eval: false

ggplot(boston_open, aes(x = days_open_at_yearend)) +
  # Histogram with a 7-day binwidth to represent weekly operational cycles
  geom_histogram(aes(y = after_stat(density)), binwidth = 7, fill = "lightgray", color = "white") +
  # Density overlay to highlight the overall shape and search for bimodality
  geom_density(color = "darkred", linewidth = 1) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(0, 365, by = 60)) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Age Distribution of Open Requests at Year-End",
    subtitle = "Investigating bimodality: Active recent backlog vs. stagnant long-term requests.",
    x = "Days Open (as of Dec 31, 2024)",
    y = "Density (Proportion of Open Work Orders)"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )
```

```{r}
#| label: fig-robust-binning-A
#| fig-subcap: true
#| layout-ncol: 3
#| fig-width: 4
#| fig-height: 4
#| echo: true
#| eval: false
#| warning: false
#| fig-cap: "Histograms of age distribution of open requests at the end of 2024, with different bin sizes: (a) days, (b) weeks and (c) months."

# Base plot for the first two charts mapping continuous days
p_base <- ggplot(boston_open, aes(x = days_open_at_yearend)) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(0, 365, by = 120)) +
  labs(x = "Days Open (as of Dec 31)", y = "Volume of Open Tickets") +
  theme(plot.title = element_text(size = 11, face = "bold"))
# Daily Binning
p_base + 
  geom_histogram(binwidth = 1, fill = "darkgray", color = "white", linewidth = 0.1) +
  labs(title = "Daily Resolution (1-Day Bins)")
# Weekly Binning
p_base + 
  geom_histogram(binwidth = 7, fill = "steelblue", color = "white", linewidth = 0.2) +
  labs(title = "Weekly Resolution (7-Day Bins)")
# Calendar Months
boston_open %>%
  # Reverse the month levels so December (newest) is on the left, matching the 0-365 days flow
  mutate(open_month_rev = factor(open_month, levels = rev(levels(open_month)))) %>%
  ggplot(aes(x = open_month_rev)) +
  geom_bar(fill = "darkred", color = "white", linewidth = 0.2) +
  theme_minimal() +
  labs(
    title = "Monthly Resolution (Calendar)",
    x = "Month Opened (Newest to Oldest)", 
    y = "Volume of Open Tickets"
  ) +
  theme(
    plot.title = element_text(size = 11, face = "bold"),
    # Angle the text slightly to ensure all 12 month labels fit cleanly
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
#| label: hartigans-dip-test-A
#| echo: true
#| warning: false
#| message: false
#| results: hold
#| eval: false

library(diptest)
library(dplyr)

# Test on the continuous days open variable
dip_result <- dip.test(boston_open$days_open_at_yearend)
# Extract and format the results
dip_stat <- round(dip_result$statistic, 4)
dip_pval <- format(dip_result$p.value, scientific = TRUE, digits = 3)
# Consolidate output into a single string
conclusion_text <- if(dip_result$p.value < 0.05) {
  "Conclusion: The p-value approaches zero, mathematically rejecting unimodality.\nThe distinct waves in the backlog are structural realities, not binning artifacts."
} else {
  "Conclusion: Failed to reject unimodality."
}

final_output <- paste0(
  "--- Hartigan's Dip Test for Unimodality ---\n",
  "Dip Statistic (D): ", dip_stat, "\n",
  "P-Value:           ", dip_pval, "\n\n",
  conclusion_text, "\n"
)
cat(final_output)
```

```{r}
#| label: fig-open-tickets-by-month-A
#| echo: true
#| eval: false
#| warning: false
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Visualization of departments contributions to open requests at the end of 2024 by starting month (Note that the month order is the reverse of that in @fig-robust-binning (c))."

# Identify the top 4 departments with the most open tickets
top_open_subjects <- boston_open %>%
  count(subject, sort = TRUE) %>%
  slice_head(n = 4) %>%
  pull(subject)
# Group the data for the stacked visualization
plot_data_open <- boston_open %>%
  mutate(
    subject_grouped = ifelse(subject %in% top_open_subjects, subject, "All Other Departments"),
    # Factor to ensure the grey baseline sits at the bottom of the stack
    subject_grouped = factor(subject_grouped, levels = c("All Other Departments", top_open_subjects))
  )
# Generate the bar chart
ggplot(plot_data_open, aes(x = open_month, fill = subject_grouped)) +
  geom_bar(color = "white", linewidth = 0.2) +
  scale_fill_manual(
    values = c(
      "All Other Departments" = "grey80",
      # Using a clear, distinct color palette for the top culprits
      setNames(c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3"), top_open_subjects)
    )
  ) +
  theme_minimal() +
  labs(
    title = "Year-End Backlog by Month of Origin",
    subtitle = "Exploratory view: Identifying seasonal waves and departmental bottlenecks.",
    x = "Month Ticket Was Opened",
    y = "Volume of Open Tickets at Year-End",
    fill = "Department"
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 9)
  ) +
  guides(fill = guide_legend(ncol = 2, reverse = TRUE))
```

## Section 4

```{r}
#| label: resolution-density-log-A
#| echo: true
#| eval: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

library(scales) 

# Summary stats 
res_summary <- boston_clean %>%
  summarise(
    Mean = mean(resolution_days, na.rm = TRUE),
    Median = median(resolution_days, na.rm = TRUE)
  )
ggplot(boston_clean, aes(x = resolution_days)) +
  geom_density(fill = "steelblue", color = "darkblue", alpha = 0.5) +
  geom_vline(xintercept = res_summary$Mean, color = "darkred", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = res_summary$Median, color = "black", linetype = "solid", linewidth = 1) +
  # log1p transformation safely handles 0-day resolutions
  scale_x_continuous(
    trans = "log1p",
    breaks = c(0, 1, 3, 10, 30, 100, 365),
    labels = label_comma()
  ) +
  # Setting y = Inf and adjusting vjust pins the labels to the top of the plot
  annotate("text", x = res_summary$Mean, y = Inf, 
           label = paste("Mean:\n", round(res_summary$Mean, 1), "days"), 
           color = "darkred", hjust = -0.2, vjust = 1.5, size = 4, fontface = "bold") +
  annotate("text", x = res_summary$Median, y = Inf, 
           label = paste("Median:\n", round(res_summary$Median, 1), "days"), 
           color = "black", hjust = 1.2, vjust = 1.5, size = 4, fontface = "bold") +
  theme_minimal() +
  labs(
    title = "Density of Maintenance Resolution Times (Log Scale)",
    subtitle = "A log1p scale stretches the immediate closures while revealing the true span of the multi-month right tail.",
    x = "Days to Close (Log Scale)",
    y = "Density of Work Orders"
  )
```

## Section 5

```{r}
#| label: fig-sla-delta-distribution-A
#| echo: true
#| eval: false
#| warning: false
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "distribution of SLA Delta."

# Calculate the exact delta in days
sla_delta_data <- boston_clean %>%
  filter(!is.na(sla_target_dt), subject != "Animal Control") %>%
  mutate(
    sla_delta_days = as.numeric(difftime(closed_dt, sla_target_dt, units = "days")),
    # Categorize for the plot fill
    compliance_status = ifelse(sla_delta_days <= 0, "On Time (Early)", "Overdue (Late)")
  )
# Plot the distribution of the delta
ggplot(sla_delta_data, aes(x = sla_delta_days, fill = compliance_status)) +
  # boundary = 0 added here to prevent overlapping bins at zero
  geom_histogram(binwidth = 1, boundary = 0, color = "white", alpha = 0.8) +
  # Add the hard SLA deadline reference line
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("On Time (Early)" = "steelblue", "Overdue (Late)" = "darkred")) +
  # Zoom the x-axis to a relevant operational window (e.g., 30 days early to 30 days late)
  coord_cartesian(xlim = c(-30, 30)) +
  theme_minimal() +
  labs(
    title = "Distribution of SLA Delta",
    subtitle = "The variance of actual closure times relative to their expected deadlines.",
    x = "SLA Delta (Days)",
    y = "Volume of Work Orders",
    fill = "Status"
  ) +
  theme(legend.position = "top")
```

```{r}
#| label: fig-ks-statistic-explanation-A
#| echo: true
#| eval: false
#| warning: false
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Graphical explanation of the Kolmogorov-Smirnov statistic."

# Identify a "Goldilocks" subgroup (high volume, moderate D_k around 0.35)
global_dist <- sla_delta_data$sla_delta_days
illustrative_group <- sla_delta_data %>%
  group_by(type) %>%
  filter(n() > 5000) %>% 
  summarise(
    ks_stat = suppressWarnings(ks.test(sla_delta_days, global_dist)$statistic),
    .groups = "drop"
  ) %>%
  mutate(diff_from_target = abs(ks_stat - 0.35)) %>%
  arrange(diff_from_target) %>%
  slice(1)
max_type_name <- illustrative_group$type
max_type_dist <- sla_delta_data %>% filter(type == max_type_name) %>% pull(sla_delta_days)
# Calculate the exact point of maximum vertical distance
x_vals <- sort(unique(c(global_dist, max_type_dist)))
ecdf_global_vals <- ecdf(global_dist)(x_vals)
ecdf_max_vals <- ecdf(max_type_dist)(x_vals)
diffs <- abs(ecdf_global_vals - ecdf_max_vals)
max_idx <- which.max(diffs)
x_max_diff <- x_vals[max_idx]
y_global <- ecdf_global_vals[max_idx]
y_max <- ecdf_max_vals[max_idx]
max_d_stat <- diffs[max_idx]
# Define dynamic labels to map math notation to the data
global_label <- "F_global : Global Baseline"
subgroup_label <- paste0("F_k : '", max_type_name, "' (Factor: type)")
# Create the dataset for plotting
plot_data <- sla_delta_data %>%
  mutate(Curve = global_label) %>%
  bind_rows(
    sla_delta_data %>%
      filter(type == max_type_name) %>%
      mutate(Curve = subgroup_label)
  )
# Generate the explanatory plot
ggplot(plot_data, aes(x = sla_delta_days, color = Curve)) +
  stat_ecdf(linewidth = 1) +
  # Add the vertical segment illustrating D_k
  annotate("segment", x = x_max_diff, xend = x_max_diff,
           y = y_global, yend = y_max,
           color = "darkred", linewidth = 1.2, 
           arrow = arrow(ends = "both", length = unit(0.1, "inches"))) +
  # Annotate the exact D_k value mathematically
  annotate("text", x = x_max_diff + 2, y = mean(c(y_global, y_max)),
           label = paste0("D[k] == ", round(max_d_stat, 3)), parse = TRUE,
           color = "darkred", hjust = 0, size = 5, fontface = "bold") +
  # Reference line for the SLA deadline (0 days)
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", alpha = 0.5) +
  # Map colors directly to the new dynamic labels
  scale_color_manual(values = setNames(c("darkgray", "steelblue"), c(global_label, subgroup_label))) +
  # Lock the zoom back to the critical operational window
  coord_cartesian(xlim = c(-30, 30)) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  labs(
    title = "Visualizing the Kolmogorov-Smirnov Statistic",
    subtitle = "The D_k statistic represents the maximum absolute vertical distance between the global distribution\nand a specific subgroup (in this example, the 'Parking Enforcement' level of the 'type' factor).",
    x = "SLA Delta (Days)",
    y = "Cumulative Proportion of Tickets Closed",
    color = NULL
  ) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 11, face = "bold")
  )
```

```{r}
#| label: feature-dependency-ks-A
#| echo: true
#| eval: false
#| warning: false
#| message: false

library(purrr)
library(dplyr)

# Global baseline distribution
global_dist <- sla_delta_data$sla_delta_days
total_n <- nrow(sla_delta_data)
factors_to_test <- c("subject", "reason", "type", "queue", "department", 
                     "fire_district", "pwd_district", "city_council_district", 
                     "police_district", "neighborhood", "neighborhood_services_district", 
                     "ward", "precinct", "location_zipcode")
# Function to calculate the Weighted K-S Distance and extract p-values
calc_weighted_ks <- function(factor_name, data, global_dist, total_n) {
  data %>%
    group_split(!!sym(factor_name)) %>%
    map_dfr(function(df) {
      level_dist <- df$sla_delta_days
      level_n <- nrow(df)
      # Skip virtually empty levels to prevent test errors
      if(level_n < 5) return(tibble(level_n = level_n, ks_stat = 0, p_val = 1))
      # Suppress warnings for expected ties
      ks_res <- suppressWarnings(ks.test(level_dist, global_dist))
      tibble(
        level_n = level_n,
        ks_stat = ks_res$statistic,
        p_val = ks_res$p.value
      )
    }) %>%
    summarise(
      Factor = factor_name,
      Weighted_KS_Score = sum(ks_stat * (level_n / total_n)),
      Median_P_Value = median(p_val)
    )
}
# Map across all factors and rank
ks_results <- map_dfr(factors_to_test, ~calc_weighted_ks(.x, sla_delta_data, global_dist, total_n)) %>%
  arrange(desc(Weighted_KS_Score)) %>%
  mutate(
    Weighted_KS_Score = round(Weighted_KS_Score, 4),
    Median_P_Value = format(Median_P_Value, scientific = TRUE, digits = 3)
  )
```

```{r}
#| label: factor-ranking-A
#| echo: true
#| eval: false
#| warning: false

library(knitr)

# Generate the formal Quarto table from the previous ks_results tibble
kable(
  ks_results, 
  col.names = c("Factor", "Weighted K-S Score", "Median P-Value"),
  align = c('l', 'c', 'c')
)
```

```{r}
#| label: fig-queue-histogram-A
#| echo: true
#| eval: false
#| warning: false
#| fig-width: 9
#| fig-height: 7

# Isolate the top 4 queues to compare against the global background
top_queues <- sla_delta_data %>%
  count(queue, sort = TRUE) %>%
  slice_head(n = 4) %>%
  pull(queue)
# Group all remaining data into a global baseline category
plot_data_queue <- sla_delta_data %>%
  mutate(
    queue_grouped = ifelse(queue %in% top_queues, queue, "All Other Queues"),
    # Convert to factor and put "All Other Queues" first so it renders at the bottom of the stack
    queue_grouped = factor(queue_grouped, levels = c("All Other Queues", top_queues))
  )
# Generate the stacked histogram
ggplot(plot_data_queue, aes(x = sla_delta_days, fill = queue_grouped)) +
  geom_histogram(binwidth = 1, boundary = 0, position = "stack", color = "white", linewidth = 0.2, alpha = 0.9) +
  # Add the SLA deadline reference line
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 1) +
  # Zoom tightly to the central mass for higher resolution
  coord_cartesian(xlim = c(-15, 15)) +
  # Manually assign colors: neutral grey for the baseline, vibrant colors for the top 4
  scale_fill_manual(
    values = c(
      "All Other Queues" = "grey80",
      setNames(c("#E41A1C", "#377EB8", "#4DAF4A", "#FF7F00"), top_queues)
    )
  ) +
  theme_minimal() +
  labs(
    title = "SLA Delta Histogram by Top Queues",
    subtitle = "Visualizing the K-S finding: Top queues form distinct distributions against the global baseline.",
    x = "SLA Delta (Days)",
    y = "Volume of Work Orders",
    fill = "Digital Queue"
  ) +
  theme(
    # Move legend to the bottom to allow more horizontal breathing room
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 9)
  ) +
  # Force the legend to wrap into 2 columns so long department names are never cut off
  guides(fill = guide_legend(ncol = 2, reverse = TRUE))
```

## Section 6

```{python}
#| label: bertopic-pipeline
#| eval: false
#| echo: true

import pandas as pd
from bertopic import BERTopic
from sentence_transformers import SentenceTransformer

# Load the SLA-filtered dataset
df = pd.read_csv("../data/boston_clean.csv") 
df = df.dropna(subset=['closure_reason'])
docs = df['closure_reason'].tolist()
# Initialize the Embedding Model on Apple Silicon GPU
embedding_model = SentenceTransformer("all-MiniLM-L6-v2", device="mps")
# Fit BERTopic to discover operational clusters 
topic_model = BERTopic(
    embedding_model=embedding_model,
    min_topic_size=50, 
    calculate_probabilities=False
)
topics, probs = topic_model.fit_transform(docs)
# Merge the AI-generated Topic Names back to the dataset
df['topic_id'] = topics
topic_info = topic_model.get_topic_info()
df = df.merge(topic_info[['Topic', 'Name']], left_on='topic_id', right_on='Topic', how='left')
df = df.rename(columns={'Name': 'ai_topic_name'})
# Export for reporting
df.to_csv("../data/boston_bertopic_results.csv", index=False)
```

```{r}
#| label: plot-bertopic-results-A
#| echo: true
#| eval: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Volume of SLA breaches categorized by AI-discovered operational themes."

library(readr)
library(dplyr)
library(ggplot2)
library(stringr)
library(here)

# Read ONLY the newly generated AI columns and the Ticket ID
ai_results <- read_csv(here("data", "boston_bertopic_results.csv")) %>%
  select(case_enquiry_id, topic_id, ai_topic_name) 
# Join the AI topics to your existing SLA dataset and filter
ai_plot_data <- sla_delta_data %>%
  inner_join(ai_results, by = "case_enquiry_id") %>%
  filter(compliance_status == "Overdue (Late)", topic_id != -1) %>%
  mutate(
    clean_topic = str_remove(ai_topic_name, "^\\d+_"),
    clean_topic = str_replace_all(clean_topic, "_", " "),
    clean_topic = str_to_title(clean_topic)
  ) %>%
  count(clean_topic, sort = TRUE) %>%
  slice_head(n = 15)
# Generate the visualization
ggplot(ai_plot_data, aes(x = n, y = reorder(clean_topic, n))) +
  geom_col(fill = "steelblue", color = "white", alpha = 0.9) +
  theme_minimal() +
  labs(
    title = "Top Causes of Operational Delays (Discovered via AI)",
    subtitle = "Unsupervised clustering of closure notes reveals distinct systemic bottlenecks.",
    x = "Volume of Delayed Work Orders",
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.y = element_text(size = 11, face = "bold")
  )
```

```{r}
#| label: llm-topic-translation-A
#| echo: true
#| eval: false
#| warning: false
#| message: false
#| cache: true

library(dplyr)
library(purrr)
library(stringr)
library(readr)
library(here)
library(knitr)
library(httr)

# CLEAN FIRST, then get the Top 15 Unique topics
ai_results <- read_csv(here("data", "boston_bertopic_results.csv"))
topic_summary <- ai_results %>%
  filter(topic_id != -1) %>%
  mutate(
    clean_raw_topic = str_remove(ai_topic_name, "^\\d+_"),
    clean_raw_topic = str_replace_all(clean_raw_topic, "_", " ")
  ) %>%
  count(clean_raw_topic, sort = TRUE) %>%
  slice_head(n = 15)
# Direct Local API with the Proper Persona
translate_topic <- function(messy_string) {
  sys_prompt <- "You are an expert municipal data analyst for the City of Boston. You classify messy 311 work order keywords."
  user_prompt <- paste0(
    "Translate this messy keyword string into a clean, 3-to-4 word operational category label (e.g., 'Equipment Maintenance', 'Duplicate Request').\n\n",
    "Return ONLY the label. Do not include quotes, formatting, or introductory text.\n\n",
    "Keyword: ", messy_string
  )
  # Send offline request directly to local Mac memory
  res <- tryCatch({
    POST(
      url = "http://localhost:11434/api/generate",
      body = list(
        model = "llama3",
        prompt = user_prompt,
        system = sys_prompt,
        stream = FALSE
      ),
      encode = "json",
      timeout(30)
    )
  }, error = function(e) return(NULL))
  if (is.null(res) || status_code(res) != 200) return("API Connection Failed")
  raw_text <- content(res, as = "parsed", type = "application/json")$response
  # Strip out any markdown or quotes the LLM tries to sneak in
  clean_label <- str_remove_all(raw_text, '["*]') %>% str_trim()
  return(as.character(clean_label))
}
# Loop through the 15 topics
llm_clean_names <- map_chr(topic_summary$clean_raw_topic, translate_topic)
# Build and print the crosswalk table
topic_crosswalk <- tibble(
  `Original c-TF-IDF String` = topic_summary$clean_raw_topic,
  `LLM Dashboard Label` = llm_clean_names
)
kable(topic_crosswalk, caption = "Automated Topic Translation via Local Llama 3 (Offline API)")
```

```{r}
#| label: fig-topic-factor-enrichment-A
#| echo: true
#| eval: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Enrichment heatmap showing which AI topics are statistically over-represented across municipal departments. Department abbreviations: BTDT (Transportation), INFO (Mayor's Hotline), ISD (Inspectional Services), PARK (Parks & Recreation), PROP (Property Management), PWDx (Public Works)."

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(here)

# Load the AI results and join the LLM translated labels
ai_df <- read_csv(here("data", "boston_bertopic_results.csv")) %>%
  filter(topic_id != -1, !is.na(department)) %>%
  mutate(
    clean_raw_topic = str_remove(ai_topic_name, "^\\d+_"),
    clean_raw_topic = str_replace_all(clean_raw_topic, "_", " ")
  ) %>%
  # Bring in the translations from the previous chunk
  left_join(topic_crosswalk, by = c("clean_raw_topic" = "Original c-TF-IDF String")) %>%
  mutate(
    # Fallback to the title-cased raw topic if it wasn't in our Top 15 translated list
    Final_Label = ifelse(is.na(`LLM Dashboard Label`), str_to_title(clean_raw_topic), `LLM Dashboard Label`)
  )
# Focus on top departments and top topics to keep the grid readable
top_depts  <- ai_df %>% count(department, sort = TRUE) %>% slice_head(n = 6) %>% pull(department)
top_topics <- ai_df %>% count(Final_Label, sort = TRUE) %>% slice_head(n = 12) %>% pull(Final_Label)
filtered_df <- ai_df %>%
  filter(department %in% top_depts, Final_Label %in% top_topics)
# Create a contingency table and run Chi-Square to get Standardized Residuals
contingency_matrix <- table(filtered_df$Final_Label, filtered_df$department)
chisq_res <- chisq.test(contingency_matrix)
# Convert residuals back to a tidy dataframe for ggplot
enrichment_data <- as.data.frame(chisq_res$stdres) %>%
  rename(Topic = Var1, Department = Var2, StdResidual = Freq)
# Visualize as an Enrichment Heatmap using the clean LLM labels
ggplot(enrichment_data, aes(x = Department, y = Topic, fill = StdResidual)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(
    low = "#2b83ba", mid = "#ffffbf", high = "#d7191c",
    midpoint = 0,
    name = "Standardized\nResidual"
  ) +
  theme_minimal() +
  labs(
    title = "Statistical Enrichment of AI Topics Across Departments",
    subtitle = "Red indicates strong over-representation (residual > +2); Blue indicates under-representation.",
    x = NULL,
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(size = 11, face = "bold")
  )
```

## Section 7

```{r}
#| label: fig-seasonal-trends-global-A
#| echo: true
#| eval: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Citywide monthly inbound ticket demand versus the volume of resulting SLA breaches."

library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(scales)

# Base dataset with extracted month
seasonal_base <- sla_delta_data %>%
  mutate(created_month = month(open_dt, label = TRUE, abbr = TRUE))
# Global Aggregation
global_summary <- seasonal_base %>%
  group_by(created_month) %>%
  summarise(
    Total_Demand = n(),
    SLA_Breaches = sum(compliance_status == "Overdue (Late)", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Total_Demand, SLA_Breaches),
    names_to = "Metric",
    values_to = "Volume"
  ) %>%
  mutate(
    Metric = factor(Metric, levels = c("Total_Demand", "SLA_Breaches"), 
                    labels = c("Total Inbound Tickets (Demand)", "Missed SLAs (System Failure)"))
  )
# Build the global seasonal overlay
ggplot(global_summary, aes(x = created_month, y = Volume, group = Metric, color = Metric)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3) +
  scale_y_continuous(labels = comma) +
  scale_color_manual(values = c("Total Inbound Tickets (Demand)" = "#b3cde3", 
                                "Missed SLAs (System Failure)" = "#e41a1c")) +
  theme_minimal() +
  labs(
    title = "Citywide Seasonal Operational Stress Test",
    subtitle = "Tracking when inbound municipal demand outpaces workforce resolution capacity.",
    x = "Month Request Opened",
    y = "Ticket Volume",
    color = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "top",
    axis.text.x = element_text(face = "bold")
  )
```

```{r}
#| label: fig-seasonal-trends-department-A
#| echo: true
#| eval: false
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Departmental breakdown of monthly demand versus SLA breaches. Department abbreviations: BTDT (Transportation), INFO (Mayor's Hotline), ISD (Inspectional Services), PARK (Parks & Recreation), PROP (Property Management), PWDx (Public Works)."

# Find the top 6 departments by overall ticket volume 
top_depts_season <- seasonal_base %>%
  count(department, sort = TRUE) %>%
  slice_head(n = 6) %>%
  pull(department)
# Departmental Aggregation
dept_summary <- seasonal_base %>%
  filter(department %in% top_depts_season) %>%
  group_by(department, created_month) %>%
  summarise(
    Total_Demand = n(),
    SLA_Breaches = sum(compliance_status == "Overdue (Late)", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Total_Demand, SLA_Breaches),
    names_to = "Metric",
    values_to = "Volume"
  ) %>%
  mutate(
    Metric = factor(Metric, levels = c("Total_Demand", "SLA_Breaches"), 
                    labels = c("Total Inbound Tickets (Demand)", "Missed SLAs (System Failure)"))
  )
# Build the faceted visualization
ggplot(dept_summary, aes(x = created_month, y = Volume, group = Metric, color = Metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  facet_wrap(~ department, scales = "free_y", ncol = 3) +
  scale_y_continuous(labels = comma) +
  scale_color_manual(values = c("Total Inbound Tickets (Demand)" = "#b3cde3", 
                                "Missed SLAs (System Failure)" = "#e41a1c")) +
  theme_minimal() +
  labs(
    title = "Seasonal Stress Test by Top Departments",
    subtitle = "Comparing capacity thresholds across individual workflows (Note: Y-axes vary by department).",
    x = "Month Request Opened",
    y = "Ticket Volume",
    color = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11),
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Section 8

```{r}
#| label: fig-survival-analysis-departments-A
#| echo: true
#| eval: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Kaplan-Meier survival curves estimating the probability a ticket remains open over time. Department abbreviations: BTDT (Transportation), INFO (Mayor's Hotline), ISD (Inspectional Services), PARK (Parks & Recreation), PROP (Property Management), PWDx (Public Works)."

library(dplyr)
library(ggplot2)
library(survival)
library(broom)
library(stringr)
library(scales)

# Isolate the top 5 departments
top_5_depts <- sla_delta_data %>%
  count(department, sort = TRUE) %>%
  slice_head(n = 5) %>%
  pull(department)
survival_df <- sla_delta_data %>%
  filter(department %in% top_5_depts) %>%
  mutate(
    max_date = max(open_dt, na.rm = TRUE),
    calc_end_dt = if_else(is.na(closed_dt), max_date, closed_dt),
    duration_days = as.numeric(difftime(calc_end_dt, open_dt, units = "days")),
    status_ind = if_else(is.na(closed_dt), 0, 1)
  ) %>%
  filter(duration_days >= 0)
# Fit the Kaplan-Meier Model
km_fit <- survfit(Surv(duration_days, status_ind) ~ department, data = survival_df)
tidy_km <- tidy(km_fit) %>% mutate(clean_dept = str_remove(strata, "department="))
# Visualize the Hazard Curves
ggplot(tidy_km, aes(x = time, y = estimate, color = clean_dept)) +
  geom_step(linewidth = 1.2, alpha = 0.8) +
  coord_cartesian(xlim = c(0, 60)) + 
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  labs(
    title = "Time-to-Resolution: Ticket Survival Analysis",
    subtitle = "The mathematical probability that a 311 request remains unresolved as days pass.",
    x = "Days Since Request Opened",
    y = "Probability Ticket is Still Open",
    color = "Department"
  ) +
  theme(plot.title = element_text(face = "bold", size = 14))
```

```{r}
#| label: fig-cox-regression-model-adjusted-A
#| echo: true
#| eval: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Hazard Ratios (Log Scale) isolating the impact of reporting sources and departments. Automated API integrations (e.g., 'Maximo Integration' used by the Boston 311 system) have been removed to isolate human workforce efficiency."

library(dplyr)
library(survival)
library(broom)
library(ggplot2)
library(forcats)
library(stringr)

# Prepare Data and Filter Automated Artifacts
cox_df <- sla_delta_data %>%
  filter(department %in% c("BTDT", "ISD", "INFO", "PARK", "PWDx")) %>%
  mutate(
    calc_end_dt = if_else(is.na(closed_dt), max(open_dt, na.rm = TRUE), closed_dt),
    duration_days = as.numeric(difftime(calc_end_dt, open_dt, units = "days")),
    status_ind = if_else(is.na(closed_dt), 0, 1),
    source_clean = if_else(is.na(source), "Unknown", as.character(source))
  ) %>%
  # Remove the Maximo API artifact to measure actual human work
  filter(duration_days > 0, !str_detect(source_clean, "Maximo"))
# Fit the Adjusted Cox Model
cox_model <- coxph(Surv(duration_days, status_ind) ~ department + source_clean, data = cox_df)
# Tidy and Exponentiate
tidy_cox <- tidy(cox_model, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "") %>%
  mutate(
    clean_term = str_remove_all(term, "department|source_clean"),
    is_significant = if_else(p.value < 0.05, "Significant (p < .05)", "Not Significant"),
    clean_term = fct_reorder(clean_term, estimate)
  )
# Build the Log-Scaled Forest Plot
ggplot(tidy_cox, aes(x = estimate, y = clean_term, color = is_significant)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black", linewidth = 1) +
  geom_point(size = 4) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, linewidth = 1.2) +
  # Apply base-10 log scale for mathematically accurate visual distances
  scale_x_log10(breaks = c(0.25, 0.5, 0.75, 1, 1.5, 2, 3, 5)) + 
  scale_color_manual(values = c("Significant (p < .05)" = "#1b9e77", "Not Significant" = "#7570b3")) +
  theme_minimal() +
  labs(
    title = "Multivariate Drivers of Ticket Resolution",
    subtitle = "Variables to the right (HR > 1) accelerate resolution. Variables to the left (HR < 1) delay it.",
    x = "Hazard Ratio (Log-Scaled Multiplier of Resolution Speed)",
    y = "Operational Variable",
    color = "Statistical Significance"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom",
    axis.text.y = element_text(face = "bold", size = 11)
  )
```

## Section 9

```{r}
#| label: fig-queue-velocity-analysis-A
#| echo: true
#| eval: false
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Queueing Stability: Weekly net change in unresolved tickets (Arrivals minus Services).Department abbreviations: BTDT (Transportation), INFO (Mayor's Hotline), ISD (Inspectional Services), PARK (Parks & Recreation), PROP (Property Management), PWDx (Public Works)."

library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)

# Calculate the Arrival Rate (lambda) and Service Rate (mu) per week
queue_dynamics <- sla_delta_data %>%
  filter(department %in% c("BTDT", "ISD", "INFO", "PARK", "PWDx")) %>%
  mutate(
    open_week = floor_date(open_dt, "week"),
    close_week = floor_date(closed_dt, "week")
  )
# Arrivals per week
arrivals <- queue_dynamics %>%
  group_by(department, week = open_week) %>%
  summarise(tickets_opened = n(), .groups = "drop")
# Services per week (ignoring NAs for open tickets)
services <- queue_dynamics %>%
  filter(!is.na(close_week)) %>%
  group_by(department, week = close_week) %>%
  summarise(tickets_closed = n(), .groups = "drop")
# Merge and calculate Net Backlog Velocity
backlog_velocity <- arrivals %>%
  full_join(services, by = c("department", "week")) %>%
  # Replace NAs with 0 for weeks where no tickets were opened or closed
  replace_na(list(tickets_opened = 0, tickets_closed = 0)) %>%
  mutate(
    net_queue_change = tickets_opened - tickets_closed,
    # Create a flag for color coding: Red = Growing queue, Blue = Shrinking queue
    status = if_else(net_queue_change > 0, "Queue Growing (Instability)", "Queue Recovering (Stability)")
  ) %>%
  # Filter out extreme edge dates if necessary to keep the plot clean
  filter(week >= min(arrivals$week, na.rm = TRUE) & week <= max(arrivals$week, na.rm = TRUE))
# Visualize the System Stability
ggplot(backlog_velocity, aes(x = week, y = net_queue_change, fill = status)) +
  geom_col(alpha = 0.85) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  facet_wrap(~ department, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("Queue Growing (Instability)" = "#e41a1c", 
                               "Queue Recovering (Stability)" = "#377eb8")) +
  theme_minimal() +
  labs(
    title = "Departmental Queue Stability (Backlog Velocity)",
    subtitle = "Tracking weekly thermodynamic balance: Tickets Opened vs. Tickets Closed.",
    x = "Week",
    y = "Net Change in Open Tickets",
    fill = "System Status"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Section 10

```{r}
#| label: fig-spatial-equity-morans-i-A
#| echo: true
#| eval: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Geospatial distribution of SLA breach rates across Boston census tracts."

library(dplyr)
library(sf)
library(tigris)
library(spdep)
library(ggplot2)
library(stringr)

# Gets the Shapefile via API (Natively downloads Suffolk County Census Tracts)
options(tigris_use_cache = TRUE)
boston_tracts <- tracts(state = "MA", county = "Suffolk", cb = TRUE, year = 2022, progress_bar = FALSE) %>%
  st_transform(crs = 4326)
# Prepare the 311 Spatial Points
spatial_311 <- sla_delta_data %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  filter(!str_detect(if_else(is.na(source), "Unknown", as.character(source)), "Maximo")) %>%
  mutate(is_breach = if_else(on_time == "ONTIME", 0, 1)) %>% 
  # Convert flat data frame into an active spatial object
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
# Spatial Join: Assign tickets to Census Tracts and calculate breach rates
tract_breaches <- boston_tracts %>%
  st_join(spatial_311, left = FALSE) %>%
  group_by(GEOID) %>%
  summarise(
    total_tickets = n(),
    breach_rate = mean(is_breach, na.rm = TRUE)
  ) %>%
  # Filter out tracts with highly volatile low sample sizes (e.g., airport, harbor islands)
  filter(total_tickets > 50)
# Calculate Global Moran's I for Spatial Autocorrelation
# Create the spatial weights matrix (defining which tracts touch each other)
neighbors <- poly2nb(tract_breaches, queen = TRUE)
weights <- nb2listw(neighbors, style = "W", zero.policy = TRUE)
# Execute the mathematical test
moran_result <- moran.test(tract_breaches$breach_rate, weights, zero.policy = TRUE)
# Visualization
ggplot(tract_breaches) +
  geom_sf(aes(fill = breach_rate), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "magma", labels = scales::percent_format()) +
  theme_void() +
  labs(
    title = "Spatial Equity of Municipal Services",
    subtitle = paste0("Global Moran's I: ", round(moran_result$estimate[1], 3), 
                      " (p-value: ", signif(moran_result$p.value, 3), ")"),
    fill = "SLA Breach Rate",
    caption = "Excludes automated Maximo Integration closures to isolate human workforce equity."
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "right"
  )
```

```{r}
#| label: fig-spatial-equity-lisa-A
#| echo: true
#| eval: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "LISA Cluster Map isolating statistically significant zones of systemic neglect (High-High) and exceptional service (Low-Low)."
# Calculate Local Moran's I
local_m <- localmoran(tract_breaches$breach_rate, weights, zero.policy = TRUE)
# Scale the variables to determine the quadrants (centered on the mean)
c_breach <- tract_breaches$breach_rate - mean(tract_breaches$breach_rate, na.rm = TRUE)
c_lag <- lag.listw(weights, c_breach, zero.policy = TRUE)
# Categorize the tracts into LISA clusters
lisa_tracts <- tract_breaches %>%
  mutate(
    # Extract the p-value from the local moran matrix (Column 5)
    p_val = local_m[, 5], 
    # Define the raw spatial quadrants
    quadrant = case_when(
      c_breach > 0 & c_lag > 0 ~ "High-High (Systemic Neglect)",
      c_breach < 0 & c_lag < 0 ~ "Low-Low (Exceptional Service)",
      c_breach > 0 & c_lag < 0 ~ "High-Low (Spatial Outlier)",
      c_breach < 0 & c_lag > 0 ~ "Low-High (Spatial Outlier)",
      TRUE ~ "Not Significant"
    ),
    # Mask the quadrants by statistical significance (p < 0.05)
    lisa_cluster = if_else(p_val <= 0.05, quadrant, "Not Significant")
  ) %>%
  # Convert the cluster to a factor for ordered plotting
  mutate(lisa_cluster = factor(lisa_cluster, levels = c(
    "High-High (Systemic Neglect)", 
    "Low-Low (Exceptional Service)", 
    "High-Low (Spatial Outlier)", 
    "Low-High (Spatial Outlier)", 
    "Not Significant"
  )))
# Map the LISA Clusters
# Define a stark, color-blind friendly palette to highlight the inequity
lisa_colors <- c(
  "High-High (Systemic Neglect)" = "#d73027",    # Red
  "Low-Low (Exceptional Service)" = "#4575b4",   # Blue
  "High-Low (Spatial Outlier)" = "#fdae61",      # Light Orange
  "Low-High (Spatial Outlier)" = "#abd9e9",      # Light Blue
  "Not Significant" = "#f0f0f0"                  # Light Grey
)
ggplot(lisa_tracts) +
  geom_sf(aes(fill = lisa_cluster), color = "darkgrey", size = 0.1) +
  scale_fill_manual(values = lisa_colors, drop = FALSE) +
  theme_void() +
  labs(
    title = "Zones of Systemic Neglect vs. Exceptional Service",
    subtitle = "Statistically significant spatial clusters of 311 SLA breaches (Local Moran's I, p < 0.05).",
    fill = "LISA Classification",
    caption = "Excludes automated 'maximo integration' closures to isolate human workforce equity."
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(face = "bold")
  )
```

