---
title: "Data Analytics for Boston 311, year 2024"
author: "Joël R. Pradines, Ph.D."
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    math: true
    embed-resources: true
    link-external-newwindow: true
bibliography: references.bib
nocite: |
  @*
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(readr)
library(ggplot2)
library(dplyr)
library(lubridate)
library(ggplot2)
library(here)
library(tidyr)
```

```{r}
#| label: ingest
#| message: false
#| warning: false
#| echo: false

boston_311 <- read_csv(here("data", "tmpm461rr5o.csv"))
```

# 1. Introduction

The introduction is about providing an example of what values descriptive analysis of a maintenance related data set can bring. In there should be the following points:
- it requires knowledge of statistics and understanding of programming
- statistical rigor is key to obtaining sound results and it requires appropriate training and experience
- analysis cannot be fully automated because each dataset is unique
- dataset uniqueness requires human detective work to uncover the dataset lessons and this often implies a trial-error process

This process is next illustrated with the analysis of the Boston 311 Year 2024 dataset which is freely available online. The detective work yields the following findings:
- TBC

# 2. Data Cleaning

Data cleaning is an important first task when analyzing a dataset. We first have to identify the format of each column, check for formatting errors and also missing data. This process is typically done via programming and it sometimes might include a few manual steps.

```{r}
#| label: data-dictionary
#| message: false
#| warning: false
#| echo: true

library(dplyr)
library(DT)

# 1. Print the overall dimensions
cat("Dataset Dimensions:", format(nrow(boston_311), big.mark = ","), "rows and", ncol(boston_311), "columns.\n\n")

# 2. Generate the column summary dataframe
data_summary <- tibble(
  Column = names(boston_311),
  Type = sapply(boston_311, function(x) paste(class(x), collapse = ", ")),
  Unique_Values = sapply(boston_311, n_distinct),
  Missing_Values = sapply(boston_311, function(x) sum(is.na(x)))
)

# 3. Render as an interactive HTML table
datatable(
  data_summary,
  rownames = FALSE,
  options = list(
    pageLength = 10, 
    autoWidth = TRUE,
    dom = 'tp' # Keeps the interface clean (just table and pagination)
  ),
  caption = "Dataset Overview: Column types, unique values, and missing data."
)
```

After validating the column formats, we filter out any requests that remained open at the end of 2024, keeping only fully resolved work orders as our primary analytical dataset.

```{r}
#| label: data-cleaning
#| message: false
#| warning: false
#| echo: true

# Ingestion
boston_311 <- read_csv(here("data", "tmpm461rr5o.csv"))
# Cleaning
boston_clean <- boston_311 %>%
  # Completed work orders
  filter(case_status == "Closed", !is.na(closed_dt), !is.na(open_dt)) %>%
  mutate(
    # Exact resolution time
    resolution_days = as.numeric(difftime(closed_dt, open_dt, units = "days")),
    # Temporal features for seasonality and scheduling analysis
    open_month = month(open_dt, label = TRUE, abbr = TRUE),
    open_wday = wday(open_dt, label = TRUE, abbr = TRUE, week_start = 1), # Monday = 1
    open_hour = hour(open_dt)
  ) %>%
  # Remove data entry errors (negative resolution times)
  filter(resolution_days >= 0)
write_csv(boston_clean, here("data", "boston_clean.csv"))
# Sort output
cat("Cleaned dataset contains", format(nrow(boston_clean), big.mark = ","), "closed records ready for analysis.\n")
```

Before proceeding to the analysis of the main dataset, it is instructive to look at some features of the subset of open requests.

# 3. Open Requests at the End of 2024

We now look at the features of the subset of open requests. This will provide an example of how to stress-test a visual observation and detect a major operational disturbance.


```{r}
#| label: isolate-open-cases
#| message: false
#| warning: false
#| echo: true

# Tickets still open at the end of 2024
boston_open <- boston_311 %>%
  filter(case_status == "Open" | is.na(closed_dt)) %>%
  mutate(
    open_month = month(open_dt, label = TRUE, abbr = TRUE),
    # Ticket age as of the end of the 2024 calendar year
    days_open_at_yearend = as.numeric(difftime(as.POSIXct("2024-12-31 23:59:59"), open_dt, units = "days"))
  )
write_csv(boston_open, here("data", "boston_open.csv"))
# Short output
cat("Isolated", format(nrow(boston_open), big.mark = ","), "open records.\n\n")
```

**Note**: *The R code was displayed ("echoed") so far to illustrate the need for programming. For the rest of the report, the code will no longer be displayed in the report output (html). But it is still contained in the source file (qmd document) that generates the html.*

@fig-open-tickets-distribution below overlays a density curve on top of a weekly-binned histogram. 

```{r}
#| label: fig-open-tickets-distribution
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Age distribution of open requests at the end of 2024."

ggplot(boston_open, aes(x = days_open_at_yearend)) +
  # Histogram with a 7-day binwidth to represent weekly operational cycles
  geom_histogram(aes(y = after_stat(density)), binwidth = 7, fill = "lightgray", color = "white") +
  # Density overlay to highlight the overall shape and search for bimodality
  geom_density(color = "darkred", linewidth = 1) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(0, 365, by = 60)) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Age Distribution of Open Requests at Year-End",
    subtitle = "Investigating bimodality: Active recent backlog vs. stagnant long-term requests.",
    x = "Days Open (as of Dec 31, 2024)",
    y = "Density (Proportion of Open Work Orders)"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )
```

Visual inspection suggests distinct "waves" or peaks. A critical step is questioning whether these structures represent real operational phenomena (like a seasonal surge) or are merely statistical illusions caused by "binning bias" — the arbitrary width of the histogram buckets.

Because our sample size is large ($N > 40,000$), we can test the structural robustness of the backlog by altering the resolution of our histogram. If the waves are real, they will persist whether we group the data by day, by week, or by month.

```{r}
#| label: fig-robust-binning
#| fig-subcap: true
#| layout-ncol: 3
#| fig-width: 4
#| fig-height: 4
#| echo: false
#| warning: false
#| fig-cap: "Histograms of age distribution of open requests at the end of 2024, with different bin sizes: (a) days, (b) weeks and (c) months."

# Base plot for the first two charts mapping continuous days
p_base <- ggplot(boston_open, aes(x = days_open_at_yearend)) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(0, 365, by = 120)) +
  labs(x = "Days Open (as of Dec 31)", y = "Volume of Open Tickets") +
  theme(plot.title = element_text(size = 11, face = "bold"))

# 1. Daily Binning
p_base + 
  geom_histogram(binwidth = 1, fill = "darkgray", color = "white", linewidth = 0.1) +
  labs(title = "Daily Resolution (1-Day Bins)")

# 2. Operational Resolution: Weekly Binning
p_base + 
  geom_histogram(binwidth = 7, fill = "steelblue", color = "white", linewidth = 0.2) +
  labs(title = "Weekly Resolution (7-Day Bins)")

# 3. Macro Resolution: Calendar Months
boston_open %>%
  # Reverse the month levels so December (newest) is on the left, matching the 0-365 days flow
  mutate(open_month_rev = factor(open_month, levels = rev(levels(open_month)))) %>%
  ggplot(aes(x = open_month_rev)) +
  geom_bar(fill = "darkred", color = "white", linewidth = 0.2) +
  theme_minimal() +
  labs(
    title = "Monthly Resolution (Calendar)",
    x = "Month Opened (Newest to Oldest)", 
    y = "Volume of Open Tickets"
  ) +
  theme(
    plot.title = element_text(size = 11, face = "bold"),
    # Angle the text slightly to ensure all 12 month labels fit cleanly
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

@fig-robust-binning suggests that the peaks at age 60-day and 170-day are robust. They do not seem to be binning artifacts. Yet, this statement is still based on visual inspection. There exists a statistically rigorous test that can be used to check for peak presence.

Hartigan's Dip Test is a non-parametric test specifically designed to detect multimodality [@hartigan1985dip]. The null hypothesis ($H_0$) states that the distribution is unimodal. Mathematically, it calculates the "dip" statistic ($D$), which is the maximum vertical distance between the empirical distribution function, $F_n(x)$, and the best-fitting unimodal distribution function, $U(x)$:
$$D = \sup_x \vert{}F_n(x) - U(x)\vert{}$$
If the data truly contains multiple structural waves (like your 60-day and 170-day spikes), $F_n(x)$ will deviate so drastically from any possible unimodal curve that the $p$-value will approach zero, definitively rejecting the null hypothesis.


```{r}
#| label: hartigans-dip-test
#| echo: false
#| warning: false
#| message: false

# Note: Requires the 'diptest' package
library(diptest)
library(dplyr)

# Execute the test on the continuous days open variable
dip_result <- dip.test(boston_open$days_open_at_yearend)

# Extract and format the results
dip_stat <- round(dip_result$statistic, 4)
dip_pval <- format(dip_result$p.value, scientific = TRUE, digits = 3)

cat("--- Hartigan's Dip Test for Unimodality ---\n")
cat("Dip Statistic (D):", dip_stat, "\n")
cat("P-Value:          ", dip_pval, "\n")

# Interpretative output for the report
if(dip_result$p.value < 0.05) {
  cat("\nConclusion: The p-value approaches zero, mathematically rejecting unimodality.\n")
  cat("The distinct waves in the backlog are structural realities, not binning artifacts.")
} else {
  cat("\nConclusion: Failed to reject unimodality.")
}
```

Having established the robustness of the 60-day and 170-day spikes, we investigate their potential origins by examining monthly requests that are still open at the end of 2024 by department in @fig-open-tickets-by-month.

```{r}
#| label: fig-open-tickets-by-month
#| echo: false
#| warning: false
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Visualization of departments contributions to open requests at the end of 2024 by starting month (Note that the month order is the reverse of that in @fig-robust-binning (c))."

# 1. Identify the top 4 departments struggling with the most open tickets
top_open_subjects <- boston_open %>%
  count(subject, sort = TRUE) %>%
  slice_head(n = 4) %>%
  pull(subject)

# 2. Group the data for the stacked visualization
plot_data_open <- boston_open %>%
  mutate(
    subject_grouped = ifelse(subject %in% top_open_subjects, subject, "All Other Departments"),
    # Factor to ensure the grey baseline sits at the bottom of the stack
    subject_grouped = factor(subject_grouped, levels = c("All Other Departments", top_open_subjects))
  )

# 3. Generate the exploratory bar chart
ggplot(plot_data_open, aes(x = open_month, fill = subject_grouped)) +
  geom_bar(color = "white", linewidth = 0.2) +
  scale_fill_manual(
    values = c(
      "All Other Departments" = "grey80",
      # Using a clear, distinct color palette for the top culprits
      setNames(c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3"), top_open_subjects)
    )
  ) +
  theme_minimal() +
  labs(
    title = "Year-End Backlog by Month of Origin",
    subtitle = "Exploratory view: Identifying seasonal waves and departmental bottlenecks.",
    x = "Month Ticket Was Opened",
    y = "Volume of Open Tickets at Year-End",
    fill = "Department"
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 9)
  ) +
  guides(fill = guide_legend(ncol = 2, reverse = TRUE))
```


The visualization immediately resolves the mystery of the 170-day spike. The massive red block shows that the Transportation - Traffic Division backlog suddenly ballooned in June 2024 and remained saturated.

This systemic blockage aligns perfectly with the timeline of the **Sumner Tunnel Restoration Project**. The Massachusetts Department of Transportation (MassDOT) executed a full, 24/7 closure of the Sumner Tunnel from July 5 to August 5, 2024. The entire month of June involved intense logistical preparation for this $160 million infrastructure project—altering traffic patterns, setting up regional detours, and modifying transit routes. This monumental shift in city traffic overwhelmed the Traffic Division's capacity, creating a localized work-order backlog that the department could not clear before the end of the year.



# 4. Resolution Time

We now focus on request resolution time. A common pitfall is relying on the average (mean) time to close a ticket. In property maintenance, completion times are rarely normally distributed. They exhibit a severe right skew: most tasks are completed quickly, but a handful of capital-intensive or disputed tasks can remain open for months. 

When a long tail exists, the mean is artificially inflated, making the operational baseline look much worse than it actually is. The median is the mathematically sound metric for setting Service Level Agreements (SLAs).


```{r}
#| label: resolution-density-log
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

library(scales) # For elegant axis breaks and labels

# Calculate the portfolio-level summary stats (if not already in a previous chunk)
res_summary <- boston_clean %>%
  summarise(
    Mean = mean(resolution_days, na.rm = TRUE),
    Median = median(resolution_days, na.rm = TRUE)
  )

ggplot(boston_clean, aes(x = resolution_days)) +
  geom_density(fill = "steelblue", color = "darkblue", alpha = 0.5) +
  geom_vline(xintercept = res_summary$Mean, color = "darkred", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = res_summary$Median, color = "black", linetype = "solid", linewidth = 1) +
  # log1p transformation safely handles 0-day resolutions
  scale_x_continuous(
    trans = "log1p",
    breaks = c(0, 1, 3, 10, 30, 100, 365),
    labels = label_comma()
  ) +
  # Setting y = Inf and adjusting vjust pins the labels to the top of the plot
  annotate("text", x = res_summary$Mean, y = Inf, 
           label = paste("Mean:\n", round(res_summary$Mean, 1), "days"), 
           color = "darkred", hjust = -0.2, vjust = 1.5, size = 4, fontface = "bold") +
  annotate("text", x = res_summary$Median, y = Inf, 
           label = paste("Median:\n", round(res_summary$Median, 1), "days"), 
           color = "black", hjust = 1.2, vjust = 1.5, size = 4, fontface = "bold") +
  theme_minimal() +
  labs(
    title = "Density of Maintenance Resolution Times (Log Scale)",
    subtitle = "A log1p scale stretches the immediate closures while revealing the true span of the multi-month right tail.",
    x = "Days to Close (Log Scale)",
    y = "Density of Work Orders"
  )
```

# 5. SLA Compliance Analytics

Absolute resolution times only tell half the story. To truly evaluate vendor and operational performance, we must compare the completion timestamp (`closed_dt`) against the algorithmic deadline (`sla_target_dt`). 

This section isolates the subset of closed requests with defined service level agreements. We calculate a compliance rate (the percentage of tickets closed on or before their target date) grouped by the top service categories.

## 5.1 The SLA Delta Distribution (Lateness/Earliness)

We now analyze the distribution of the SLA Delta (Actual Resolution Time minus SLA Target Time). 
A value of `0` means the ticket was closed exactly on the deadline. Negative values indicate early closures, while positive values indicate the ticket was overdue. Looking at this continuous distribution reveals the variance and behavioral clustering around the deadline.

```{r}
#| label: sla-delta-distribution
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

# Calculate the exact delta in days
sla_delta_data <- boston_clean %>%
  filter(!is.na(sla_target_dt), subject != "Animal Control") %>%
  mutate(
    sla_delta_days = as.numeric(difftime(closed_dt, sla_target_dt, units = "days")),
    # Categorize for the plot fill
    compliance_status = ifelse(sla_delta_days <= 0, "On Time (Early)", "Overdue (Late)")
  )

# Plot the distribution of the delta
ggplot(sla_delta_data, aes(x = sla_delta_days, fill = compliance_status)) +
  # boundary = 0 added here to prevent overlapping bins at zero
  geom_histogram(binwidth = 1, boundary = 0, color = "white", alpha = 0.8) +
  # Add the hard SLA deadline reference line
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("On Time (Early)" = "steelblue", "Overdue (Late)" = "darkred")) +
  # Zoom the x-axis to a relevant operational window (e.g., 30 days early to 30 days late)
  coord_cartesian(xlim = c(-30, 30)) +
  theme_minimal() +
  labs(
    title = "Distribution of SLA Delta",
    subtitle = "The variance of actual closure times relative to their expected deadlines.",
    x = "SLA Delta (Days)",
    y = "Volume of Work Orders",
    fill = "Status"
  ) +
  theme(legend.position = "top")
```


## 5.2 Factors Driving Variance in SLA Compliance

To systematically identify which factors drive early or late resolutions, we test the dependency between our categorical features (e.g., department, queue, geography,...) and the continuous SLA Delta. 

Because the dataset contains right-tail outliers, traditional variance-based tests (like an ANOVA) are not statistically sound to apply. In addition, the large $N$ value guarantees that all $p$-values approach zero, and the extreme skew breaks the underlying assumptions of ordinary least squares (OLS), rendering variance explained ($R^2$) unreliable.

### 5.2.1 The Mathematical Approach: Volume-Weighted K-S Distance

To correctly capture the true operational reality, we abandon variance and compare the **Empirical Cumulative Distribution Functions (ECDFs)** using a custom Volume-Weighted Kolmogorov-Smirnov (K-S) Statistic.


First, we define $F_{global}(x)$ as the ECDF of the SLA Delta for the entire city-wide portfolio. For any given factor (e.g., Department), we split the data into $K$ individual levels (subgroups). We then define $F_k(x)$ as the ECDF for a specific subgroup $k$.

The standard K-S statistic $D_k$ measures the maximum absolute vertical distance between the subgroup's ECDF and the global ECDF:

$$D_k = \sup_x |F_k(x) - F_{global}(x)|$$

This outputs a bounded value between $0$ (identical distributions) and $1$ (zero overlap). Because it relies purely on rank and percentiles, it is called a nonparametric statistic and it is immune to extreme outliers.

```{r}
#| label: fig-ks-statistic-explanation
#| echo: false
#| warning: false
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Graphical explanation of the Kolmogorov-Smirnov statistic."
# 1. Identify a "Goldilocks" subgroup (high volume, moderate D_k around 0.35)
global_dist <- sla_delta_data$sla_delta_days

illustrative_group <- sla_delta_data %>%
  group_by(type) %>%
  filter(n() > 5000) %>% 
  summarise(
    ks_stat = suppressWarnings(ks.test(sla_delta_days, global_dist)$statistic),
    .groups = "drop"
  ) %>%
  mutate(diff_from_target = abs(ks_stat - 0.35)) %>%
  arrange(diff_from_target) %>%
  slice(1)

max_type_name <- illustrative_group$type
max_type_dist <- sla_delta_data %>% filter(type == max_type_name) %>% pull(sla_delta_days)

# 2. Calculate the exact point of maximum vertical distance
x_vals <- sort(unique(c(global_dist, max_type_dist)))
ecdf_global_vals <- ecdf(global_dist)(x_vals)
ecdf_max_vals <- ecdf(max_type_dist)(x_vals)

diffs <- abs(ecdf_global_vals - ecdf_max_vals)
max_idx <- which.max(diffs)

x_max_diff <- x_vals[max_idx]
y_global <- ecdf_global_vals[max_idx]
y_max <- ecdf_max_vals[max_idx]
max_d_stat <- diffs[max_idx]

# Define dynamic labels to map math notation to the data
global_label <- "F_global : Global Baseline"
subgroup_label <- paste0("F_k : '", max_type_name, "' (Factor: type)")

# 3. Create the dataset for plotting
plot_data <- sla_delta_data %>%
  mutate(Curve = global_label) %>%
  bind_rows(
    sla_delta_data %>%
      filter(type == max_type_name) %>%
      mutate(Curve = subgroup_label)
  )

# 4. Generate the explanatory plot
ggplot(plot_data, aes(x = sla_delta_days, color = Curve)) +
  stat_ecdf(linewidth = 1) +
  # Add the vertical segment illustrating D_k
  annotate("segment", x = x_max_diff, xend = x_max_diff,
           y = y_global, yend = y_max,
           color = "darkred", linewidth = 1.2, 
           arrow = arrow(ends = "both", length = unit(0.1, "inches"))) +
  # Annotate the exact D_k value mathematically
  annotate("text", x = x_max_diff + 2, y = mean(c(y_global, y_max)),
           label = paste0("D[k] == ", round(max_d_stat, 3)), parse = TRUE,
           color = "darkred", hjust = 0, size = 5, fontface = "bold") +
  # Reference line for the SLA deadline (0 days)
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", alpha = 0.5) +
  # Map colors directly to the new dynamic labels
  scale_color_manual(values = setNames(c("darkgray", "steelblue"), c(global_label, subgroup_label))) +
  # Lock the zoom back to the critical operational window
  coord_cartesian(xlim = c(-30, 30)) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  labs(
    title = "Visualizing the Kolmogorov-Smirnov Statistic",
    subtitle = "The D_k statistic represents the maximum absolute vertical distance between the global distribution\nand a specific subgroup (in this example, the 'Parking Enforcement' level of the 'type' factor).",
    x = "SLA Delta (Days)",
    y = "Cumulative Proportion of Tickets Closed",
    color = NULL
  ) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 11, face = "bold")
  )
```

To evaluate the *overall* explanatory power of the categorical factor, we must aggregate the $D_k$ scores of its levels. A simple average would mathematically equate a queue with $10$ tickets to a queue with $10,000$ tickets. Therefore, we calculate the **Volume-Weighted K-S Score** ($D_{weighted}$) by weighting each subgroup's K-S statistic by its proportional volume:

$$D_{weighted} = \sum_{k=1}^{K} \left( \frac{n_k}{N} \right) D_k$$

Where $n_k$ is the total number of work orders in subgroup $k$, and $N$ is the total volume across all subgroups. This penalizes noisy variance from tiny subgroups and rewards factors where massive subsets of work fundamentally deviate from the global baseline.

### 5.2.2 Results: Task vs. Geography

When we apply this non-parametric ranking to all organizational, task-based, and geographic features, a stark operational divide emerges:

* **Task & Routing Govern Compliance:** The top five drivers of distributional shift (`type`, `queue`, `reason`, `subject`, and `department`) are entirely organizational. The physical nature of the request and the specific digital bucket it is routed to fundamentally alter the timeline.
* **Geographic Equity:** Spatial features (`precinct`, `ward`, `zipcode`, and municipal districts) rank uniformly at the bottom. This mathematically proves geographic equity; a property's location does not dictate the operational urgency or speed of its service.

```{r}
#| label: feature-dependency-ks
#| echo: false
#| warning: false
#| message: false

library(purrr)
library(dplyr)

# Isolate the global baseline distribution
global_dist <- sla_delta_data$sla_delta_days
total_n <- nrow(sla_delta_data)

factors_to_test <- c("subject", "reason", "type", "queue", "department", 
                     "fire_district", "pwd_district", "city_council_district", 
                     "police_district", "neighborhood", "neighborhood_services_district", 
                     "ward", "precinct", "location_zipcode")

# Function to calculate the Weighted K-S Distance and extract p-values
calc_weighted_ks <- function(factor_name, data, global_dist, total_n) {
  
  data %>%
    group_split(!!sym(factor_name)) %>%
    map_dfr(function(df) {
      level_dist <- df$sla_delta_days
      level_n <- nrow(df)
      
      # Skip virtually empty levels to prevent test errors
      if(level_n < 5) return(tibble(level_n = level_n, ks_stat = 0, p_val = 1))
      
      # Suppress warnings for expected ties
      ks_res <- suppressWarnings(ks.test(level_dist, global_dist))
      
      tibble(
        level_n = level_n,
        ks_stat = ks_res$statistic,
        p_val = ks_res$p.value
      )
    }) %>%
    summarise(
      Factor = factor_name,
      Weighted_KS_Score = sum(ks_stat * (level_n / total_n)),
      Median_P_Value = median(p_val)
    )
}

# Map across all factors and rank
ks_results <- map_dfr(factors_to_test, ~calc_weighted_ks(.x, sla_delta_data, global_dist, total_n)) %>%
  arrange(desc(Weighted_KS_Score)) %>%
  mutate(
    Weighted_KS_Score = round(Weighted_KS_Score, 4),
    Median_P_Value = format(Median_P_Value, scientific = TRUE, digits = 3)
  )
```

```{r}
#| label: tbl-factor-ranking
#| tbl-cap: "Factor Dependency Ranking (Weighted K-S Distance)"
#| echo: false
#| warning: false

library(knitr)

# Generate the formal Quarto table from our previous ks_results tibble
kable(
  ks_results, 
  col.names = c("Factor", "Weighted K-S Score", "Median P-Value"),
  align = c('l', 'c', 'c')
)
```

### 5.2.3 Result Illustration with Histograms

While the K-S statistic is mathematically bulletproof against outliers, it is an abstract metric. To ground this finding back, we can look directly at the raw data for one of the highest-scoring factors: queue (D-score = 0.373).

By coloring the overall SLA Delta histogram by the top digital queues, we visually validate what the K-S test just measured. The distinct colors do not overlap uniformly; rather, they form separate, distinct operational distributions spanning from early completions to massive late backlogs.

```{r}
#| label: fig-queue-histogram
#| echo: false
#| warning: false
#| fig-width: 9
#| fig-height: 7

# 1. Isolate the top 4 queues to compare against the global background
top_queues <- sla_delta_data %>%
  count(queue, sort = TRUE) %>%
  slice_head(n = 4) %>%
  pull(queue)

# 2. Group all remaining data into a global baseline category
plot_data_queue <- sla_delta_data %>%
  mutate(
    queue_grouped = ifelse(queue %in% top_queues, queue, "All Other Queues"),
    # Convert to factor and put "All Other Queues" first so it renders at the bottom of the stack
    queue_grouped = factor(queue_grouped, levels = c("All Other Queues", top_queues))
  )

# 3. Generate the stacked histogram
ggplot(plot_data_queue, aes(x = sla_delta_days, fill = queue_grouped)) +
  geom_histogram(binwidth = 1, boundary = 0, position = "stack", color = "white", linewidth = 0.2, alpha = 0.9) +
  # Add the SLA deadline reference line
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 1) +
  # Zoom tightly to the central mass for higher resolution
  coord_cartesian(xlim = c(-15, 15)) +
  # Manually assign colors: neutral grey for the baseline, vibrant colors for the top 4
  scale_fill_manual(
    values = c(
      "All Other Queues" = "grey80",
      setNames(c("#E41A1C", "#377EB8", "#4DAF4A", "#FF7F00"), top_queues)
    )
  ) +
  theme_minimal() +
  labs(
    title = "SLA Delta Histogram by Top Queues",
    subtitle = "Visualizing the K-S finding: Top queues form distinct distributions against the global baseline.",
    x = "SLA Delta (Days)",
    y = "Volume of Work Orders",
    fill = "Digital Queue"
  ) +
  theme(
    # Move legend to the bottom to allow more horizontal breathing room
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 9)
  ) +
  # Force the legend to wrap into 2 columns so long department names are never cut off
  guides(fill = guide_legend(ncol = 2, reverse = TRUE))
```

# 6. Unsupervised Text Classification (BERTopic)

To uncover the specific operational themes driving SLA delays without relying on human bias or arbitrary keyword guessing, we deployed an unsupervised deep learning pipeline. 

Using the `sentence-transformers` architecture locally on Apple Silicon, we converted the unstructured closure notes into high-dimensional embeddings, which were then mathematically clustered using HDBSCAN. Finally, the algorithm generated human-readable topic labels for these clusters.

Because Transformer models require heavy computation, the classification script below was executed asynchronously. The results were saved to a static dataset, ensuring this report remains lightweight and strictly focused on descriptive visualization.

### 6.1 The Python AI Pipeline

```python
#| label: bertopic-pipeline
#| eval: false
#| echo: true

import pandas as pd
from bertopic import BERTopic
from sentence_transformers import SentenceTransformer

# 1. Load the SLA-filtered dataset
df = pd.read_csv("../data/boston_clean.csv") 
df = df.dropna(subset=['closure_reason'])
docs = df['closure_reason'].tolist()

# 2. Initialize the Embedding Model on Apple Silicon GPU
embedding_model = SentenceTransformer("all-MiniLM-L6-v2", device="mps")

# 3. Fit BERTopic to discover operational clusters 
topic_model = BERTopic(
    embedding_model=embedding_model,
    min_topic_size=50, 
    calculate_probabilities=False
)
topics, probs = topic_model.fit_transform(docs)

# 4. Merge the AI-generated Topic Names back to the dataset
df['topic_id'] = topics
topic_info = topic_model.get_topic_info()
df = df.merge(topic_info[['Topic', 'Name']], left_on='topic_id', right_on='Topic', how='left')
df = df.rename(columns={'Name': 'ai_topic_name'})

# 5. Export for reporting
df.to_csv("../data/boston_bertopic_results.csv", index=False)
```

### 6.2 Visualizing the AI Clusters

With the AI having categorized the unstructured text, we can now visualize the exact volume of delayed tickets attributed to each systemic issue. (Note: BERTopic assigns a topic ID of -1 to statistical outliers—tickets with highly unique text that do not fit into any broader operational pattern).

```{r}
#| label: plot-bertopic-results
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Volume of SLA breaches categorized by AI-discovered operational themes."

library(readr)
library(dplyr)
library(ggplot2)
library(stringr)
library(here)

# 1. Read ONLY the newly generated AI columns and the Ticket ID
ai_results <- read_csv(here("data", "boston_bertopic_results.csv")) %>%
  select(case_enquiry_id, topic_id, ai_topic_name) 

# 2. Join the AI topics to your existing SLA dataset and filter
ai_plot_data <- sla_delta_data %>%
  inner_join(ai_results, by = "case_enquiry_id") %>%
  filter(compliance_status == "Overdue (Late)", topic_id != -1) %>%
  mutate(
    clean_topic = str_remove(ai_topic_name, "^\\d+_"),
    clean_topic = str_replace_all(clean_topic, "_", " "),
    clean_topic = str_to_title(clean_topic)
  ) %>%
  count(clean_topic, sort = TRUE) %>%
  slice_head(n = 15)

# 3. Generate the visualization
ggplot(ai_plot_data, aes(x = n, y = reorder(clean_topic, n))) +
  geom_col(fill = "steelblue", color = "white", alpha = 0.9) +
  theme_minimal() +
  labs(
    title = "Top Causes of Operational Delays (Discovered via AI)",
    subtitle = "Unsupervised clustering of closure notes reveals distinct systemic bottlenecks.",
    x = "Volume of Delayed Work Orders",
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.y = element_text(size = 11, face = "bold")
  )
```

```{r}
#| label: topic-factor-enrichment
#| echo: true
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Enrichment heatmap showing which AI topics are statistically over-represented across municipal departments."

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(here)

# 1. Load the AI results
ai_df <- read_csv(here("data", "boston_bertopic_results.csv")) %>%
  filter(topic_id != -1, !is.na(department)) %>%
  mutate(
    clean_topic = str_remove(ai_topic_name, "^\\d+_"),
    clean_topic = str_replace_all(clean_topic, "_", " "),
    clean_topic = str_to_title(clean_topic)
  )

# 2. Focus on top departments and top topics to keep the grid readable
top_depts  <- ai_df %>% count(department, sort = TRUE) %>% slice_head(n = 6) %>% pull(department)
top_topics <- ai_df %>% count(clean_topic, sort = TRUE) %>% slice_head(n = 12) %>% pull(clean_topic)

filtered_df <- ai_df %>%
  filter(department %in% top_depts, clean_topic %in% top_topics)

# 3. Create a contingency table and run Chi-Square to get Standardized Residuals
contingency_matrix <- table(filtered_df$clean_topic, filtered_df$department)
chisq_res <- chisq.test(contingency_matrix)

# 4. Convert residuals back to a tidy dataframe for ggplot
enrichment_data <- as.data.frame(chisq_res$stdres) %>%
  rename(Topic = Var1, Department = Var2, StdResidual = Freq)

# 5. Visualize as an Enrichment Heatmap
ggplot(enrichment_data, aes(x = Department, y = Topic, fill = StdResidual)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(
    low = "#2b83ba", mid = "#ffffbf", high = "#d7191c",
    midpoint = 0,
    name = "Standardized\nResidual"
  ) +
  theme_minimal() +
  labs(
    title = "Statistical Enrichment of AI Topics Across Departments",
    subtitle = "Red indicates strong over-representation (residual > +2); Blue indicates under-representation.",
    x = NULL,
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(size = 10)
  )
```