---
title: "Data Analytics for Boston 311, year 2024"
author: "Joël R. Pradines, Ph.D."
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    math: true
    embed-resources: true
    link-external-newwindow: true
bibliography: references.bib
nocite: |
  @*
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(readr)
library(ggplot2)
library(dplyr)
library(lubridate)
library(ggplot2)
library(here)
library(tidyr)
```

```{r}
#| label: ingest
#| message: false
#| warning: false
#| echo: false

boston_311 <- read_csv(here("data", "tmpm461rr5o.csv"))
```

# 1. Introduction: Applied Analytics for Maintenance Operations

Managing large-scale property portfolios or municipal infrastructures—spanning hundreds of facilities and thousands of service requests—requires more than basic descriptive dashboards. Standard reporting metrics, such as average completion times or binary SLA (Service Level Agreement) compliance rates, frequently mask systemic bottlenecks. They obscure capital constraints, disguise inefficient dispatch routing, and fail to measure true workforce capacity.

Optimizing maintenance operations at scale requires the application of advanced Data Science (DS) and Operations Research (OR) methodologies. Extracting actionable operational intelligence from service databases is an iterative process that relies on three core pillars:

Deep Statistical Rigor: Standard averages are mathematically flawed in maintenance data due to survivorship bias and right-skewed extremes that prohibit traditional normal assumptions. Robust probabilistic modeling is required to measure true performance and system stability. This requires someone having deep understanding of and extensive experience is statistics.

Human Detective Work: Every operational dataset contains unique administrative artifacts, batch-processing illusions, and mislabeled text. Uncovering the truth requires an exploratory, trial-and-error analytical process that cannot be fully automated by generalized software.

Advanced Programming: Moving beyond spreadsheet limitations to deploy multivariate regressions, spatial autocorrelation, and localized machine learning pipelines is required to uncover valuable findings.

## The Example Dataset (Boston 311)

To demonstrate the application and analytical value of these methodologies, this report utilizes the publicly available Boston 311 (Year 2024) dataset as an applied proxy.

In this context, Boston’s geographic districts and digital routing queues serve as an excellent mathematical analog for decentralized property management. Just as a massive real estate portfolio is divided into distinct building subsets managed by local technician teams, the city's workload is distributed across distinct neighborhoods and operational queues. By treating constituent requests as tenant service orders, this document serves as a methodological blueprint, detailing step-by-step how advanced analytics can uncover local team bottlenecks, evaluate geographic equity across different property subsets, and model true workforce dynamics.


# 2. Data Validation and Engineering

Data validation forms the foundation of any rigorous statistical analysis. Before engineering operational features, we must programmatically audit the raw dataset to map column formats, quantify unique values, and identify missing records (NAs). This step prevents silent failures in downstream mathematical models.

```{r}
#| label: data-dictionary
#| message: false
#| warning: false
#| echo: true

library(dplyr)
library(DT)

# 1. Print the overall dimensions
cat("Dataset Dimensions:", format(nrow(boston_311), big.mark = ","), "rows and", ncol(boston_311), "columns.\n\n")

# 2. Generate the column summary dataframe
data_summary <- tibble(
  Column = names(boston_311),
  Type = sapply(boston_311, function(x) paste(class(x), collapse = ", ")),
  Unique_Values = sapply(boston_311, n_distinct),
  Missing_Values = sapply(boston_311, function(x) sum(is.na(x)))
)

# 3. Render as an interactive HTML table
datatable(
  data_summary,
  rownames = FALSE,
  options = list(
    pageLength = 10, 
    autoWidth = TRUE,
    dom = 'tp' # Keeps the interface clean (just table and pagination)
  ),
  caption = "Dataset Overview: Column types, unique values, and missing data."
)
```

Following the initial audit, we bifurcate the raw data into two distinct analytical subsets to isolate different operational dynamics:

The Backlog Cohort (Open Requests): Work orders that remained unresolved at the end of the observation period. Analyzing this subset is critical for diagnosing structural blockages and current system capacity.

The Resolved Cohort (Closed Requests): Work orders with a complete lifecycle. This forms our primary analytical dataset for evaluating historical workforce velocity and Service Level Agreement (SLA) compliance.

The pipeline below isolates the resolved cohort, purges data entry anomalies (such as negative resolution times), and engineers new temporal features to prepare for subsequent seasonality and scheduling analysis.

```{r}
#| label: data-cleaning
#| message: false
#| warning: false
#| echo: true

# Ingestion
boston_311 <- read_csv(here("data", "tmpm461rr5o.csv"))

# Cleaning and Feature Engineering for the Resolved Cohort
boston_clean <- boston_311 %>%
  # Filter for completed work orders
  filter(case_status == "Closed", !is.na(closed_dt), !is.na(open_dt)) %>%
  mutate(
    # Exact resolution time
    resolution_days = as.numeric(difftime(closed_dt, open_dt, units = "days")),
    # Temporal features for seasonality and scheduling analysis
    open_month = month(open_dt, label = TRUE, abbr = TRUE),
    open_wday = wday(open_dt, label = TRUE, abbr = TRUE, week_start = 1), # Monday = 1
    open_hour = hour(open_dt)
  ) %>%
  # Remove data entry errors (negative resolution times)
  filter(resolution_days >= 0)

write_csv(boston_clean, here("data", "boston_clean.csv"))

# Sort output
cat("Cleaned dataset contains", format(nrow(boston_clean), big.mark = ","), "closed records ready for analysis.\n")
```

Before evaluating historical performance with the resolved cohort, we will first analyze the backlog cohort to diagnose the current structural health of the system.

# 3. Open Requests at the End of 2024

We now look at the features of the subset of open requests. This will provide an example of how to stress-test a visual observation and detect a major operational disturbance.


```{r}
#| label: isolate-open-cases
#| message: false
#| warning: false
#| echo: true

# Tickets still open at the end of 2024
boston_open <- boston_311 %>%
  filter(case_status == "Open" | is.na(closed_dt)) %>%
  mutate(
    open_month = month(open_dt, label = TRUE, abbr = TRUE),
    # Ticket age as of the end of the 2024 calendar year
    days_open_at_yearend = as.numeric(difftime(as.POSIXct("2024-12-31 23:59:59"), open_dt, units = "days"))
  )
write_csv(boston_open, here("data", "boston_open.csv"))
# Short output
cat("Isolated", format(nrow(boston_open), big.mark = ","), "open records.\n\n")
```

@fig-open-tickets-distribution below overlays a density curve on top of a weekly-binned histogram. 

```{r}
#| label: fig-open-tickets-distribution
#| echo: true
#| warning: false
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Age distribution of open requests at the end of 2024."

ggplot(boston_open, aes(x = days_open_at_yearend)) +
  # Histogram with a 7-day binwidth to represent weekly operational cycles
  geom_histogram(aes(y = after_stat(density)), binwidth = 7, fill = "lightgray", color = "white") +
  # Density overlay to highlight the overall shape and search for bimodality
  geom_density(color = "darkred", linewidth = 1) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(0, 365, by = 60)) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Age Distribution of Open Requests at Year-End",
    subtitle = "Investigating bimodality: Active recent backlog vs. stagnant long-term requests.",
    x = "Days Open (as of Dec 31, 2024)",
    y = "Density (Proportion of Open Work Orders)"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )
```

Visual inspection suggests distinct "waves" or peaks. A required step is questioning whether these structures represent real operational phenomena (like a seasonal surge) or are statistical artifacts caused by "binning bias" — the chosen width of the histogram buckets.

Because our sample size is large ($N > 40,000$), we can test the structural robustness of the backlog by altering the resolution of our histogram. If the waves are real, they will persist whether we group the data by day, by week, or by month.

```{r}
#| label: fig-robust-binning
#| fig-subcap: true
#| layout-ncol: 3
#| fig-width: 4
#| fig-height: 4
#| echo: true
#| warning: false
#| fig-cap: "Histograms of age distribution of open requests at the end of 2024, with different bin sizes: (a) days, (b) weeks and (c) months."

# Base plot for the first two charts mapping continuous days
p_base <- ggplot(boston_open, aes(x = days_open_at_yearend)) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(0, 365, by = 120)) +
  labs(x = "Days Open (as of Dec 31)", y = "Volume of Open Tickets") +
  theme(plot.title = element_text(size = 11, face = "bold"))

# 1. Daily Binning
p_base + 
  geom_histogram(binwidth = 1, fill = "darkgray", color = "white", linewidth = 0.1) +
  labs(title = "Daily Resolution (1-Day Bins)")

# 2. Operational Resolution: Weekly Binning
p_base + 
  geom_histogram(binwidth = 7, fill = "steelblue", color = "white", linewidth = 0.2) +
  labs(title = "Weekly Resolution (7-Day Bins)")

# 3. Macro Resolution: Calendar Months
boston_open %>%
  # Reverse the month levels so December (newest) is on the left, matching the 0-365 days flow
  mutate(open_month_rev = factor(open_month, levels = rev(levels(open_month)))) %>%
  ggplot(aes(x = open_month_rev)) +
  geom_bar(fill = "darkred", color = "white", linewidth = 0.2) +
  theme_minimal() +
  labs(
    title = "Monthly Resolution (Calendar)",
    x = "Month Opened (Newest to Oldest)", 
    y = "Volume of Open Tickets"
  ) +
  theme(
    plot.title = element_text(size = 11, face = "bold"),
    # Angle the text slightly to ensure all 12 month labels fit cleanly
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

@fig-robust-binning suggests that the peaks at age 60-day and 170-day are robust. They do not seem to be binning artifacts. Yet, this statement is still based on visual inspection. So, the next step is to use an existing statistically rigorous test to check for peak presence.

Hartigan's Dip Test is a non-parametric test specifically designed to detect multimodality [@hartigan1985dip]. The null hypothesis ($H_0$) states that the distribution is unimodal. Mathematically, it calculates the "dip" statistic ($D$), which is the maximum vertical distance between the empirical distribution function, $F_n(x)$, and the best-fitting unimodal distribution function, $U(x)$:
$$D = \sup_x \vert{}F_n(x) - U(x)\vert{}$$
If the data truly contains multiple structural waves (like the 60-day and 170-day spikes), $F_n(x)$ will deviate so drastically from any possible unimodal curve that the $p$-value will approach zero, definitively rejecting the null hypothesis.


```{r}
#| label: hartigans-dip-test
#| echo: true
#| warning: false
#| message: false
#| results: hold

# Note: Requires the 'diptest' package
library(diptest)
library(dplyr)

# Execute the test on the continuous days open variable
dip_result <- dip.test(boston_open$days_open_at_yearend)

# Extract and format the results
dip_stat <- round(dip_result$statistic, 4)
dip_pval <- format(dip_result$p.value, scientific = TRUE, digits = 3)

# Consolidate output into a single string to ensure clean HTML rendering
conclusion_text <- if(dip_result$p.value < 0.05) {
  "Conclusion: The p-value approaches zero, mathematically rejecting unimodality.\nThe distinct waves in the backlog are structural realities, not binning artifacts."
} else {
  "Conclusion: Failed to reject unimodality."
}

final_output <- paste0(
  "--- Hartigan's Dip Test for Unimodality ---\n",
  "Dip Statistic (D): ", dip_stat, "\n",
  "P-Value:           ", dip_pval, "\n\n",
  conclusion_text, "\n"
)

# Print once
cat(final_output)
```


Having established the robustness of the 60-day and 170-day spikes, we investigate their potential origins by examining monthly requests that are still open at the end of 2024 by department in @fig-open-tickets-by-month.

```{r}
#| label: fig-open-tickets-by-month
#| echo: true
#| warning: false
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Visualization of departments contributions to open requests at the end of 2024 by starting month (Note that the month order is the reverse of that in @fig-robust-binning (c))."

# 1. Identify the top 4 departments struggling with the most open tickets
top_open_subjects <- boston_open %>%
  count(subject, sort = TRUE) %>%
  slice_head(n = 4) %>%
  pull(subject)

# 2. Group the data for the stacked visualization
plot_data_open <- boston_open %>%
  mutate(
    subject_grouped = ifelse(subject %in% top_open_subjects, subject, "All Other Departments"),
    # Factor to ensure the grey baseline sits at the bottom of the stack
    subject_grouped = factor(subject_grouped, levels = c("All Other Departments", top_open_subjects))
  )

# 3. Generate the exploratory bar chart
ggplot(plot_data_open, aes(x = open_month, fill = subject_grouped)) +
  geom_bar(color = "white", linewidth = 0.2) +
  scale_fill_manual(
    values = c(
      "All Other Departments" = "grey80",
      # Using a clear, distinct color palette for the top culprits
      setNames(c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3"), top_open_subjects)
    )
  ) +
  theme_minimal() +
  labs(
    title = "Year-End Backlog by Month of Origin",
    subtitle = "Exploratory view: Identifying seasonal waves and departmental bottlenecks.",
    x = "Month Ticket Was Opened",
    y = "Volume of Open Tickets at Year-End",
    fill = "Department"
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 9)
  ) +
  guides(fill = guide_legend(ncol = 2, reverse = TRUE))
```


The visualization resolves the mystery of the 170-day spike. The massive red block shows that the Transportation - Traffic Division backlog suddenly ballooned in June 2024 and remained saturated.

This systemic blockage aligns perfectly with the timeline of the **Sumner Tunnel Restoration Project**. The Massachusetts Department of Transportation (MassDOT) executed a full, 24/7 closure of the Sumner Tunnel from July 5 to August 5, 2024. The entire month of June involved intense logistical preparation for this $160 million infrastructure project—altering traffic patterns, setting up regional detours, and modifying transit routes. This monumental shift in city traffic overwhelmed the Traffic Division's capacity, creating a localized work-order backlog that the department could not clear before the end of the year.

While the June spike represents a clear exogenous infrastructure shock (the Sumner Tunnel), the visualization reveals a secondary wave originating in October (the ~60-day backlog)  which is not a broad seasonal shift across the city, but a second, distinct infrastructure shock affecting that exact same department: the Transportation - Traffic Division.

In October 2024, the regional transit authority (MBTA) executed a massive, multi-week shutdown of the Orange Line subway system, replacing underground trains with fleets of surface-level shuttle buses. To accommodate this, the Traffic Division had to rapidly implement severe street-level modifications across the city's southwest corridor—including removing hundreds of parking spaces, creating pop-up bus lanes, and altering one-way traffic circulations.

Just as with the summer tunnel closure, this large logistical shift overwhelmed the Traffic Division's capacity. They were forced to abandon routine neighborhood work orders to manage the crisis, mathematically cementing a massive secondary backlog into the system right before the calendar year closed.


# 4. Resolution Time

We now focus on request resolution time. A common pitfall is relying on the average (mean) time to close a ticket. In property maintenance, completion times are rarely normally distributed. They exhibit a severe right skew: most tasks are completed quickly, but a handful of capital-intensive or disputed tasks can remain open for months. 

When a long tail exists, the mean is artificially inflated, making the operational baseline look much worse than it actually is. The median is the mathematically sound metric for setting Service Level Agreements (SLAs).


```{r}
#| label: resolution-density-log
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

library(scales) # For elegant axis breaks and labels

# Calculate the portfolio-level summary stats (if not already in a previous chunk)
res_summary <- boston_clean %>%
  summarise(
    Mean = mean(resolution_days, na.rm = TRUE),
    Median = median(resolution_days, na.rm = TRUE)
  )

ggplot(boston_clean, aes(x = resolution_days)) +
  geom_density(fill = "steelblue", color = "darkblue", alpha = 0.5) +
  geom_vline(xintercept = res_summary$Mean, color = "darkred", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = res_summary$Median, color = "black", linetype = "solid", linewidth = 1) +
  # log1p transformation safely handles 0-day resolutions
  scale_x_continuous(
    trans = "log1p",
    breaks = c(0, 1, 3, 10, 30, 100, 365),
    labels = label_comma()
  ) +
  # Setting y = Inf and adjusting vjust pins the labels to the top of the plot
  annotate("text", x = res_summary$Mean, y = Inf, 
           label = paste("Mean:\n", round(res_summary$Mean, 1), "days"), 
           color = "darkred", hjust = -0.2, vjust = 1.5, size = 4, fontface = "bold") +
  annotate("text", x = res_summary$Median, y = Inf, 
           label = paste("Median:\n", round(res_summary$Median, 1), "days"), 
           color = "black", hjust = 1.2, vjust = 1.5, size = 4, fontface = "bold") +
  theme_minimal() +
  labs(
    title = "Density of Maintenance Resolution Times (Log Scale)",
    subtitle = "A log1p scale stretches the immediate closures while revealing the true span of the multi-month right tail.",
    x = "Days to Close (Log Scale)",
    y = "Density of Work Orders"
  )
```

As the density visualization illustrates, even after applying a logarithmic transformation, the distribution remains heavily right-skewed and multi-modal. The data fundamentally fails to achieve log-normality. This mathematical reality precludes the use of standard parametric statistical models (which assume a normal distribution of errors) and necessitates the advanced, non-parametric, and survival-based methodologies deployed in the subsequent sections of this report.

# 5. SLA Compliance Analytics

Absolute resolution times only tell half the story. To truly evaluate vendor and operational performance, we must compare the completion timestamp (`closed_dt`) against the algorithmic deadline (`sla_target_dt`). 

This section isolates the subset of closed requests with defined service level agreements. We calculate a compliance rate (the percentage of tickets closed on or before their target date) grouped by the top service categories.

## 5.1 The SLA Delta Distribution (Lateness/Earliness)

We now analyze the distribution of the SLA Delta (Actual Resolution Time minus SLA Target Time). 
A value of `0` means the ticket was closed exactly on the deadline. Negative values indicate early closures, while positive values indicate the ticket was overdue. Looking at this continuous distribution reveals the variance and behavioral clustering around the deadline.

```{r}
#| label: sla-delta-distribution
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

# Calculate the exact delta in days
sla_delta_data <- boston_clean %>%
  filter(!is.na(sla_target_dt), subject != "Animal Control") %>%
  mutate(
    sla_delta_days = as.numeric(difftime(closed_dt, sla_target_dt, units = "days")),
    # Categorize for the plot fill
    compliance_status = ifelse(sla_delta_days <= 0, "On Time (Early)", "Overdue (Late)")
  )

# Plot the distribution of the delta
ggplot(sla_delta_data, aes(x = sla_delta_days, fill = compliance_status)) +
  # boundary = 0 added here to prevent overlapping bins at zero
  geom_histogram(binwidth = 1, boundary = 0, color = "white", alpha = 0.8) +
  # Add the hard SLA deadline reference line
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("On Time (Early)" = "steelblue", "Overdue (Late)" = "darkred")) +
  # Zoom the x-axis to a relevant operational window (e.g., 30 days early to 30 days late)
  coord_cartesian(xlim = c(-30, 30)) +
  theme_minimal() +
  labs(
    title = "Distribution of SLA Delta",
    subtitle = "The variance of actual closure times relative to their expected deadlines.",
    x = "SLA Delta (Days)",
    y = "Volume of Work Orders",
    fill = "Status"
  ) +
  theme(legend.position = "top")
```


## 5.2 Factors Driving Variance in SLA Compliance

To systematically identify which factors drive early or late resolutions, we test the dependency between our categorical features (e.g., department, queue, geography,...) and the continuous SLA Delta. 

Because the dataset contains right-tail outliers, traditional variance-based tests (like an ANOVA) are not statistically sound to apply. In addition, the large $N$ value guarantees that all $p$-values approach zero, and the extreme skew breaks the underlying assumptions of ordinary least squares (OLS), rendering variance explained ($R^2$) unreliable.

### 5.2.1 The Mathematical Approach: Volume-Weighted K-S Distance

To correctly capture the true operational reality, we abandon variance and compare the **Empirical Cumulative Distribution Functions (ECDFs)** using a custom Volume-Weighted Kolmogorov-Smirnov (K-S) Statistic.


First, we define $F_{global}(x)$ as the ECDF of the SLA Delta for the entire city-wide portfolio. For any given factor (e.g., Department), we split the data into $K$ individual levels (subgroups). We then define $F_k(x)$ as the ECDF for a specific subgroup $k$.

The standard K-S statistic $D_k$ measures the maximum absolute vertical distance between the subgroup's ECDF and the global ECDF:

$$D_k = \sup_x |F_k(x) - F_{global}(x)|$$

This outputs a bounded value between $0$ (identical distributions) and $1$ (zero overlap). Because it relies purely on rank and percentiles, it is called a nonparametric statistic and it is immune to extreme outliers.

```{r}
#| label: fig-ks-statistic-explanation
#| echo: false
#| warning: false
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Graphical explanation of the Kolmogorov-Smirnov statistic."
# 1. Identify a "Goldilocks" subgroup (high volume, moderate D_k around 0.35)
global_dist <- sla_delta_data$sla_delta_days

illustrative_group <- sla_delta_data %>%
  group_by(type) %>%
  filter(n() > 5000) %>% 
  summarise(
    ks_stat = suppressWarnings(ks.test(sla_delta_days, global_dist)$statistic),
    .groups = "drop"
  ) %>%
  mutate(diff_from_target = abs(ks_stat - 0.35)) %>%
  arrange(diff_from_target) %>%
  slice(1)

max_type_name <- illustrative_group$type
max_type_dist <- sla_delta_data %>% filter(type == max_type_name) %>% pull(sla_delta_days)

# 2. Calculate the exact point of maximum vertical distance
x_vals <- sort(unique(c(global_dist, max_type_dist)))
ecdf_global_vals <- ecdf(global_dist)(x_vals)
ecdf_max_vals <- ecdf(max_type_dist)(x_vals)

diffs <- abs(ecdf_global_vals - ecdf_max_vals)
max_idx <- which.max(diffs)

x_max_diff <- x_vals[max_idx]
y_global <- ecdf_global_vals[max_idx]
y_max <- ecdf_max_vals[max_idx]
max_d_stat <- diffs[max_idx]

# Define dynamic labels to map math notation to the data
global_label <- "F_global : Global Baseline"
subgroup_label <- paste0("F_k : '", max_type_name, "' (Factor: type)")

# 3. Create the dataset for plotting
plot_data <- sla_delta_data %>%
  mutate(Curve = global_label) %>%
  bind_rows(
    sla_delta_data %>%
      filter(type == max_type_name) %>%
      mutate(Curve = subgroup_label)
  )

# 4. Generate the explanatory plot
ggplot(plot_data, aes(x = sla_delta_days, color = Curve)) +
  stat_ecdf(linewidth = 1) +
  # Add the vertical segment illustrating D_k
  annotate("segment", x = x_max_diff, xend = x_max_diff,
           y = y_global, yend = y_max,
           color = "darkred", linewidth = 1.2, 
           arrow = arrow(ends = "both", length = unit(0.1, "inches"))) +
  # Annotate the exact D_k value mathematically
  annotate("text", x = x_max_diff + 2, y = mean(c(y_global, y_max)),
           label = paste0("D[k] == ", round(max_d_stat, 3)), parse = TRUE,
           color = "darkred", hjust = 0, size = 5, fontface = "bold") +
  # Reference line for the SLA deadline (0 days)
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", alpha = 0.5) +
  # Map colors directly to the new dynamic labels
  scale_color_manual(values = setNames(c("darkgray", "steelblue"), c(global_label, subgroup_label))) +
  # Lock the zoom back to the critical operational window
  coord_cartesian(xlim = c(-30, 30)) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  labs(
    title = "Visualizing the Kolmogorov-Smirnov Statistic",
    subtitle = "The D_k statistic represents the maximum absolute vertical distance between the global distribution\nand a specific subgroup (in this example, the 'Parking Enforcement' level of the 'type' factor).",
    x = "SLA Delta (Days)",
    y = "Cumulative Proportion of Tickets Closed",
    color = NULL
  ) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 11, face = "bold")
  )
```

To evaluate the *overall* explanatory power of the categorical factor, we must aggregate the $D_k$ scores of its levels. A simple average would mathematically equate a queue with $10$ tickets to a queue with $10,000$ tickets. Therefore, we calculate the **Volume-Weighted K-S Score** ($D_{weighted}$) by weighting each subgroup's K-S statistic by its proportional volume:

$$D_{weighted} = \sum_{k=1}^{K} \left( \frac{n_k}{N} \right) D_k$$

Where $n_k$ is the total number of work orders in subgroup $k$, and $N$ is the total volume across all subgroups. This penalizes noisy variance from tiny subgroups and rewards factors where massive subsets of work fundamentally deviate from the global baseline.

### 5.2.2 Results: Task vs. Geography

When we apply this non-parametric ranking to all organizational, task-based, and geographic features, a stark operational divide emerges:

* **Task & Routing Govern Compliance:** The top five drivers of distributional shift (`type`, `queue`, `reason`, `subject`, and `department`) are entirely organizational. The physical nature of the request and the specific digital bucket it is routed to fundamentally alter the timeline.
* **Geographic Equity:** Spatial features (`precinct`, `ward`, `zipcode`, and municipal districts) rank uniformly at the bottom. This mathematically proves geographic equity; a property's location does not dictate the operational urgency or speed of its service.

```{r}
#| label: feature-dependency-ks
#| echo: false
#| warning: false
#| message: false

library(purrr)
library(dplyr)

# Isolate the global baseline distribution
global_dist <- sla_delta_data$sla_delta_days
total_n <- nrow(sla_delta_data)

factors_to_test <- c("subject", "reason", "type", "queue", "department", 
                     "fire_district", "pwd_district", "city_council_district", 
                     "police_district", "neighborhood", "neighborhood_services_district", 
                     "ward", "precinct", "location_zipcode")

# Function to calculate the Weighted K-S Distance and extract p-values
calc_weighted_ks <- function(factor_name, data, global_dist, total_n) {
  
  data %>%
    group_split(!!sym(factor_name)) %>%
    map_dfr(function(df) {
      level_dist <- df$sla_delta_days
      level_n <- nrow(df)
      
      # Skip virtually empty levels to prevent test errors
      if(level_n < 5) return(tibble(level_n = level_n, ks_stat = 0, p_val = 1))
      
      # Suppress warnings for expected ties
      ks_res <- suppressWarnings(ks.test(level_dist, global_dist))
      
      tibble(
        level_n = level_n,
        ks_stat = ks_res$statistic,
        p_val = ks_res$p.value
      )
    }) %>%
    summarise(
      Factor = factor_name,
      Weighted_KS_Score = sum(ks_stat * (level_n / total_n)),
      Median_P_Value = median(p_val)
    )
}

# Map across all factors and rank
ks_results <- map_dfr(factors_to_test, ~calc_weighted_ks(.x, sla_delta_data, global_dist, total_n)) %>%
  arrange(desc(Weighted_KS_Score)) %>%
  mutate(
    Weighted_KS_Score = round(Weighted_KS_Score, 4),
    Median_P_Value = format(Median_P_Value, scientific = TRUE, digits = 3)
  )
```

```{r}
#| label: tbl-factor-ranking
#| tbl-cap: "Factor Dependency Ranking (Weighted K-S Distance)"
#| echo: false
#| warning: false

library(knitr)

# Generate the formal Quarto table from our previous ks_results tibble
kable(
  ks_results, 
  col.names = c("Factor", "Weighted K-S Score", "Median P-Value"),
  align = c('l', 'c', 'c')
)
```

### 5.2.3 Result Illustration with Histograms

While the K-S statistic is mathematically bulletproof against outliers, it is an abstract metric. To ground this finding back, we can look directly at the raw data for one of the highest-scoring factors: queue (D-score = 0.373).

By coloring the overall SLA Delta histogram by the top digital queues, we visually validate what the K-S test just measured. The distinct colors do not overlap uniformly; rather, they form separate, distinct operational distributions spanning from early completions to massive late backlogs.

```{r}
#| label: fig-queue-histogram
#| echo: false
#| warning: false
#| fig-width: 9
#| fig-height: 7

# 1. Isolate the top 4 queues to compare against the global background
top_queues <- sla_delta_data %>%
  count(queue, sort = TRUE) %>%
  slice_head(n = 4) %>%
  pull(queue)

# 2. Group all remaining data into a global baseline category
plot_data_queue <- sla_delta_data %>%
  mutate(
    queue_grouped = ifelse(queue %in% top_queues, queue, "All Other Queues"),
    # Convert to factor and put "All Other Queues" first so it renders at the bottom of the stack
    queue_grouped = factor(queue_grouped, levels = c("All Other Queues", top_queues))
  )

# 3. Generate the stacked histogram
ggplot(plot_data_queue, aes(x = sla_delta_days, fill = queue_grouped)) +
  geom_histogram(binwidth = 1, boundary = 0, position = "stack", color = "white", linewidth = 0.2, alpha = 0.9) +
  # Add the SLA deadline reference line
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 1) +
  # Zoom tightly to the central mass for higher resolution
  coord_cartesian(xlim = c(-15, 15)) +
  # Manually assign colors: neutral grey for the baseline, vibrant colors for the top 4
  scale_fill_manual(
    values = c(
      "All Other Queues" = "grey80",
      setNames(c("#E41A1C", "#377EB8", "#4DAF4A", "#FF7F00"), top_queues)
    )
  ) +
  theme_minimal() +
  labs(
    title = "SLA Delta Histogram by Top Queues",
    subtitle = "Visualizing the K-S finding: Top queues form distinct distributions against the global baseline.",
    x = "SLA Delta (Days)",
    y = "Volume of Work Orders",
    fill = "Digital Queue"
  ) +
  theme(
    # Move legend to the bottom to allow more horizontal breathing room
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 9)
  ) +
  # Force the legend to wrap into 2 columns so long department names are never cut off
  guides(fill = guide_legend(ncol = 2, reverse = TRUE))
```

# 6. Unsupervised Text Classification (BERTopic & Ollama)

To uncover the specific operational themes driving SLA delays without relying on human bias or arbitrary keyword guessing, we deployed an unsupervised deep learning pipeline.

Using the sentence-transformers architecture [@reimers2019] locally on **Apple Silicon**, we converted the unstructured closure notes into high-dimensional embeddings, which were then mathematically clustered using HDBSCAN. Because the raw statistical labels generated by this process can be difficult to read, we integrated a local Large Language Model (Llama 3 via Ollama) to translate these clusters into clean, executive-ready dashboard categories.

Because Transformer models require heavy computation, the initial Python classification script below was executed asynchronously. The results were saved to a static dataset, ensuring this report remains lightweight and strictly focused on descriptive visualization.

### 6.1 The Python AI Pipeline

We utilized BERTopic [@grootendorst2022] combined with a Hugging Face embedding model to group the closure notes into distinct operational bottlenecks.

```python
#| label: bertopic-pipeline
#| eval: false
#| echo: true

import pandas as pd
from bertopic import BERTopic
from sentence_transformers import SentenceTransformer

# 1. Load the SLA-filtered dataset
df = pd.read_csv("../data/boston_clean.csv") 
df = df.dropna(subset=['closure_reason'])
docs = df['closure_reason'].tolist()

# 2. Initialize the Embedding Model on Apple Silicon GPU
embedding_model = SentenceTransformer("all-MiniLM-L6-v2", device="mps")

# 3. Fit BERTopic to discover operational clusters 
topic_model = BERTopic(
    embedding_model=embedding_model,
    min_topic_size=50, 
    calculate_probabilities=False
)
topics, probs = topic_model.fit_transform(docs)

# 4. Merge the AI-generated Topic Names back to the dataset
df['topic_id'] = topics
topic_info = topic_model.get_topic_info()
df = df.merge(topic_info[['Topic', 'Name']], left_on='topic_id', right_on='Topic', how='left')
df = df.rename(columns={'Name': 'ai_topic_name'})

# 5. Export for reporting
df.to_csv("../data/boston_bertopic_results.csv", index=False)
```

### 6.2 Visualizing the AI Clusters

With the AI having categorized the unstructured text, we can now visualize the exact volume of delayed tickets attributed to each systemic issue. We apply a lightweight text-cleaning function in R to make the raw c-TF-IDF labels readable. (Note: BERTopic assigns a topic ID of -1 to statistical outliers—tickets with highly unique text that do not fit into any broader operational pattern).

```{r}
#| label: plot-bertopic-results
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Volume of SLA breaches categorized by AI-discovered operational themes."

library(readr)
library(dplyr)
library(ggplot2)
library(stringr)
library(here)

# 1. Read ONLY the newly generated AI columns and the Ticket ID
ai_results <- read_csv(here("data", "boston_bertopic_results.csv")) %>%
  select(case_enquiry_id, topic_id, ai_topic_name) 

# 2. Join the AI topics to your existing SLA dataset and filter
ai_plot_data <- sla_delta_data %>%
  inner_join(ai_results, by = "case_enquiry_id") %>%
  filter(compliance_status == "Overdue (Late)", topic_id != -1) %>%
  mutate(
    clean_topic = str_remove(ai_topic_name, "^\\d+_"),
    clean_topic = str_replace_all(clean_topic, "_", " "),
    clean_topic = str_to_title(clean_topic)
  ) %>%
  count(clean_topic, sort = TRUE) %>%
  slice_head(n = 15)

# 3. Generate the visualization
ggplot(ai_plot_data, aes(x = n, y = reorder(clean_topic, n))) +
  geom_col(fill = "steelblue", color = "white", alpha = 0.9) +
  theme_minimal() +
  labs(
    title = "Top Causes of Operational Delays (Discovered via AI)",
    subtitle = "Unsupervised clustering of closure notes reveals distinct systemic bottlenecks.",
    x = "Volume of Delayed Work Orders",
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.y = element_text(size = 11, face = "bold")
  )
```

## 6.3 Refining Labels with Local Generative AI (Ollama)

While the plots above successfully utilize the mathematical c-TF-IDF keyword strings (e.g., "Related Responds Patrols Encourage"), these labels lack human context.

To bridge the gap between machine-readable statistics and executive-ready dashboards, we execute a 100% offline workflow using R's httr package to query a local Llama 3 instance via Ollama. This acts as an automated formatting bot, applying municipal context to translate the messy statistical outputs into polished categories. We will use these clean labels for all subsequent deep-dive analysis.

```{r}
#| label: llm-topic-translation
#| echo: true
#| warning: false
#| message: false
#| cache: true

library(dplyr)
library(purrr)
library(stringr)
library(readr)
library(here)
library(knitr)
library(httr)

# 1. CLEAN FIRST, then get the Top 15 Unique topics
ai_results <- read_csv(here("data", "boston_bertopic_results.csv"))

topic_summary <- ai_results %>%
  filter(topic_id != -1) %>%
  mutate(
    clean_raw_topic = str_remove(ai_topic_name, "^\\d+_"),
    clean_raw_topic = str_replace_all(clean_raw_topic, "_", " ")
  ) %>%
  count(clean_raw_topic, sort = TRUE) %>%
  slice_head(n = 15)

# 2. Direct Local API with the Proper Persona
translate_topic <- function(messy_string) {
  sys_prompt <- "You are an expert municipal data analyst for the City of Boston. You classify messy 311 work order keywords."
  user_prompt <- paste0(
    "Translate this messy keyword string into a clean, 3-to-4 word operational category label (e.g., 'Equipment Maintenance', 'Duplicate Request').\n\n",
    "Return ONLY the label. Do not include quotes, formatting, or introductory text.\n\n",
    "Keyword: ", messy_string
  )
  
  # Send offline request directly to local Mac memory
  res <- tryCatch({
    POST(
      url = "http://localhost:11434/api/generate",
      body = list(
        model = "llama3",
        prompt = user_prompt,
        system = sys_prompt,
        stream = FALSE
      ),
      encode = "json",
      timeout(30)
    )
  }, error = function(e) return(NULL))
  
  if (is.null(res) || status_code(res) != 200) return("API Connection Failed")
  
  raw_text <- content(res, as = "parsed", type = "application/json")$response
  
  # Strip out any markdown or quotes the LLM tries to sneak in
  clean_label <- str_remove_all(raw_text, '["*]') %>% str_trim()
  
  return(as.character(clean_label))
}

# 3. Loop through the 15 topics
llm_clean_names <- map_chr(topic_summary$clean_raw_topic, translate_topic)

# 4. Build and print the crosswalk table
topic_crosswalk <- tibble(
  `Original c-TF-IDF String` = topic_summary$clean_raw_topic,
  `LLM Dashboard Label` = llm_clean_names
)

kable(topic_crosswalk, caption = "Automated Topic Translation via Local Llama 3 (Offline API)")
```

## 6.4 Departmental Topic Enrichment

Now equipped with our clean, AI-generated operational labels, we can mathematically test how these unsupervised clusters map to actual municipal departments.

Using Standardized Pearson Residuals from a Chi-Square test, this heatmap highlights which operational bottlenecks are statistically over-represented in specific departmental workflows.

```{r}
#| label: topic-factor-enrichment
#| echo: true
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Enrichment heatmap showing which AI topics are statistically over-represented across municipal departments."

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(here)

# 1. Load the AI results and join the LLM translated labels
ai_df <- read_csv(here("data", "boston_bertopic_results.csv")) %>%
  filter(topic_id != -1, !is.na(department)) %>%
  mutate(
    clean_raw_topic = str_remove(ai_topic_name, "^\\d+_"),
    clean_raw_topic = str_replace_all(clean_raw_topic, "_", " ")
  ) %>%
  # Bring in the translations from the previous chunk
  left_join(topic_crosswalk, by = c("clean_raw_topic" = "Original c-TF-IDF String")) %>%
  mutate(
    # Fallback to the title-cased raw topic if it wasn't in our Top 15 translated list
    Final_Label = ifelse(is.na(`LLM Dashboard Label`), str_to_title(clean_raw_topic), `LLM Dashboard Label`)
  )

# 2. Focus on top departments and top topics to keep the grid readable
top_depts  <- ai_df %>% count(department, sort = TRUE) %>% slice_head(n = 6) %>% pull(department)
top_topics <- ai_df %>% count(Final_Label, sort = TRUE) %>% slice_head(n = 12) %>% pull(Final_Label)

filtered_df <- ai_df %>%
  filter(department %in% top_depts, Final_Label %in% top_topics)

# 3. Create a contingency table and run Chi-Square to get Standardized Residuals
contingency_matrix <- table(filtered_df$Final_Label, filtered_df$department)
chisq_res <- chisq.test(contingency_matrix)

# 4. Convert residuals back to a tidy dataframe for ggplot
enrichment_data <- as.data.frame(chisq_res$stdres) %>%
  rename(Topic = Var1, Department = Var2, StdResidual = Freq)

# 5. Visualize as an Enrichment Heatmap using the clean LLM labels
ggplot(enrichment_data, aes(x = Department, y = Topic, fill = StdResidual)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(
    low = "#2b83ba", mid = "#ffffbf", high = "#d7191c",
    midpoint = 0,
    name = "Standardized\nResidual"
  ) +
  theme_minimal() +
  labs(
    title = "Statistical Enrichment of AI Topics Across Departments",
    subtitle = "Red indicates strong over-representation (residual > +2); Blue indicates under-representation.",
    x = NULL,
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(size = 11, face = "bold")
  )
```

# 7. Seasonal Demand and System Capacity

To properly engineer workforce schedules, we must understand the seasonal fluctuations in constituent demand. However, high demand alone does not necessarily indicate a problem if the municipal workforce is properly scaled to handle it.

The true operational insight lies in comparing total inbound ticket volume (Demand) against the volume of missed SLAs (System Failure) month over month. This reveals the exact moments when operational capacity is overwhelmed by seasonal surges.

## 7.1 Citywide Macro Trends

We first look at the aggregate stress placed on Boston's municipal infrastructure across the entire year.

```{r}
#| label: seasonal-trends-global
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Citywide monthly inbound ticket demand versus the volume of resulting SLA breaches."

library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(scales)

# 1. Base dataset with extracted month
seasonal_base <- sla_delta_data %>%
  mutate(created_month = month(open_dt, label = TRUE, abbr = TRUE))

# 2. Global Aggregation
global_summary <- seasonal_base %>%
  group_by(created_month) %>%
  summarise(
    Total_Demand = n(),
    SLA_Breaches = sum(compliance_status == "Overdue (Late)", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Total_Demand, SLA_Breaches),
    names_to = "Metric",
    values_to = "Volume"
  ) %>%
  mutate(
    Metric = factor(Metric, levels = c("Total_Demand", "SLA_Breaches"), 
                    labels = c("Total Inbound Tickets (Demand)", "Missed SLAs (System Failure)"))
  )

# 3. Build the global seasonal overlay
ggplot(global_summary, aes(x = created_month, y = Volume, group = Metric, color = Metric)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3) +
  scale_y_continuous(labels = comma) +
  scale_color_manual(values = c("Total Inbound Tickets (Demand)" = "#b3cde3", 
                                "Missed SLAs (System Failure)" = "#e41a1c")) +
  theme_minimal() +
  labs(
    title = "Citywide Seasonal Operational Stress Test",
    subtitle = "Tracking when inbound municipal demand outpaces workforce resolution capacity.",
    x = "Month Request Opened",
    y = "Ticket Volume",
    color = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "top",
    axis.text.x = element_text(face = "bold")
  )
```

## 7.2 Departmental Micro Trends

While the citywide view provides the baseline, operational failures are rarely distributed evenly. By breaking the seasonality down by the highest-volume departments, we can see how different seasonal events (e.g., winter storms vs. summer park usage) trigger capacity thresholds across distinct municipal workforces.

```{r}
#| label: seasonal-trends-department
#| echo: false
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Departmental breakdown of monthly demand versus SLA breaches."

# 1. Find the top 6 departments by overall ticket volume 
top_depts_season <- seasonal_base %>%
  count(department, sort = TRUE) %>%
  slice_head(n = 6) %>%
  pull(department)

# 2. Departmental Aggregation
dept_summary <- seasonal_base %>%
  filter(department %in% top_depts_season) %>%
  group_by(department, created_month) %>%
  summarise(
    Total_Demand = n(),
    SLA_Breaches = sum(compliance_status == "Overdue (Late)", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Total_Demand, SLA_Breaches),
    names_to = "Metric",
    values_to = "Volume"
  ) %>%
  mutate(
    Metric = factor(Metric, levels = c("Total_Demand", "SLA_Breaches"), 
                    labels = c("Total Inbound Tickets (Demand)", "Missed SLAs (System Failure)"))
  )

# 3. Build the faceted visualization
ggplot(dept_summary, aes(x = created_month, y = Volume, group = Metric, color = Metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  facet_wrap(~ department, scales = "free_y", ncol = 3) +
  scale_y_continuous(labels = comma) +
  scale_color_manual(values = c("Total Inbound Tickets (Demand)" = "#b3cde3", 
                                "Missed SLAs (System Failure)" = "#e41a1c")) +
  theme_minimal() +
  labs(
    title = "Seasonal Stress Test by Top Departments",
    subtitle = "Comparing capacity thresholds across individual workflows (Note: Y-axes vary by department).",
    x = "Month Request Opened",
    y = "Ticket Volume",
    color = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11),
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# 8. Operational Reliability (Survival Analysis)

While average resolution times are commonly used in municipal reporting, they suffer from a severe statistical flaw: survivorship bias. By excluding tickets that are currently open, standard averages artificially deflate the apparent resolution time of heavily backlogged departments.

## 8.1 Kaplan-Meier

To capture the true operational reliability of the city, we apply time-to-event Survival Analysis. Using the Kaplan-Meier estimator, we plot the exact probability that a constituent's request will remain unresolved as days pass, handling open tickets as right-censored data points.

```{r}
#| label: survival-analysis-departments
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Kaplan-Meier survival curves estimating the probability a ticket remains open over time."

library(dplyr)
library(ggplot2)
library(survival)
library(broom)
library(stringr)
library(scales)

# 1. Isolate the top 5 departments
top_5_depts <- sla_delta_data %>%
  count(department, sort = TRUE) %>%
  slice_head(n = 5) %>%
  pull(department)

survival_df <- sla_delta_data %>%
  filter(department %in% top_5_depts) %>%
  mutate(
    max_date = max(open_dt, na.rm = TRUE),
    calc_end_dt = if_else(is.na(closed_dt), max_date, closed_dt),
    duration_days = as.numeric(difftime(calc_end_dt, open_dt, units = "days")),
    status_ind = if_else(is.na(closed_dt), 0, 1)
  ) %>%
  filter(duration_days >= 0)

# 2. Fit the Kaplan-Meier Model
km_fit <- survfit(Surv(duration_days, status_ind) ~ department, data = survival_df)
tidy_km <- tidy(km_fit) %>% mutate(clean_dept = str_remove(strata, "department="))

# 3. Visualize the Hazard Curves
ggplot(tidy_km, aes(x = time, y = estimate, color = clean_dept)) +
  geom_step(linewidth = 1.2, alpha = 0.8) +
  coord_cartesian(xlim = c(0, 60)) + 
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  labs(
    title = "Time-to-Resolution: Ticket Survival Analysis",
    subtitle = "The mathematical probability that a 311 request remains unresolved as days pass.",
    x = "Days Since Request Opened",
    y = "Probability Ticket is Still Open",
    color = "Department"
  ) +
  theme(plot.title = element_text(face = "bold", size = 14))
```

The Kaplan-Meier estimation reveals stark differences in how municipal departments process their queues, highlighting both highly reactive workforces and structural bottlenecks:

Public Works (PWDx) - The Reactive Engine: The orange curve plummets almost instantly, indicating that roughly 80% to 85% of their tickets are resolved within the first 48 to 72 hours. This mathematically depicts a workforce optimized for rapid, high-volume triage (e.g., immediate pothole patching, trash pickup).

The Baseline Cluster (BTDT, INFO, ISD): Transportation, the Mayor's Hotline, and Inspectional Services track closely together. They successfully clear about 75% of their open tickets within the first week. After that initial drop, they demonstrate a steady, predictable burn-down rate for more complex cases, approaching a near-zero unresolved rate by the 60-day mark.

Parks and Recreation (PARK) - The Structural Long-Tail: The purple curve is the critical anomaly. While PARK resolves about half of its tickets in the first few days, the curve abruptly flattens into a concrete floor. By day 20, nearly 30% of their tickets remain open, and approximately 20% are essentially abandoned past the 60-day mark. This mathematical "long-tail" confirms that PARK does not suffer from a simple speed issue, but rather a structural inability to clear complex tickets, heavily implying capital budget constraints or a reliance on seasonal workforce unlocks.

## 8.2 Multivariate Drivers of Resolution (Cox Proportional Hazards)

While the Kaplan-Meier curves show us the overall survival rate of tickets by department, municipal operations do not exist in a vacuum. A ticket's resolution time is simultaneously influenced by the department assigned, the method the constituent used to report it, and whether it includes rich context like photographs.To untangle these overlapping variables, we apply a Cox Proportional Hazards Model:$$h(t\vert{}X) = h_0(t) \exp(\beta_1 X_1 + \beta_2 X_2 + \dots + \beta_p X_p)$$This multivariate regression calculates a Hazard Ratio (HR) for each variable. In this context, a "hazard" is a positive event (closing the ticket).HR > 1: The variable mathematically accelerates resolution (e.g., an HR of 1.5 means the ticket is closed 50% faster than the baseline).HR < 1: The variable acts as a bottleneck, delaying resolution.R#| label: cox-regression-model

```{r}
#| label: cox-regression-model-adjusted
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Hazard Ratios (Log Scale) isolating the impact of reporting sources and departments. Automated API integrations have been removed to isolate human workforce efficiency."

library(dplyr)
library(survival)
library(broom)
library(ggplot2)
library(forcats)
library(stringr)

# 1. Prepare Data and Filter Automated Artifacts
cox_df <- sla_delta_data %>%
  filter(department %in% c("BTDT", "ISD", "INFO", "PARK", "PWDx")) %>%
  mutate(
    calc_end_dt = if_else(is.na(closed_dt), max(open_dt, na.rm = TRUE), closed_dt),
    duration_days = as.numeric(difftime(calc_end_dt, open_dt, units = "days")),
    status_ind = if_else(is.na(closed_dt), 0, 1),
    source_clean = if_else(is.na(source), "Unknown", as.character(source))
  ) %>%
  # Remove the Maximo API artifact to measure actual human work
  filter(duration_days > 0, !str_detect(source_clean, "Maximo"))

# 2. Fit the Adjusted Cox Model
cox_model <- coxph(Surv(duration_days, status_ind) ~ department + source_clean, data = cox_df)

# 3. Tidy and Exponentiate
tidy_cox <- tidy(cox_model, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "") %>%
  mutate(
    clean_term = str_remove_all(term, "department|source_clean"),
    is_significant = if_else(p.value < 0.05, "Significant (p < .05)", "Not Significant"),
    clean_term = fct_reorder(clean_term, estimate)
  )

# 4. Build the Log-Scaled Forest Plot
ggplot(tidy_cox, aes(x = estimate, y = clean_term, color = is_significant)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black", linewidth = 1) +
  geom_point(size = 4) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, linewidth = 1.2) +
  # Apply base-10 log scale for mathematically accurate visual distances
  scale_x_log10(breaks = c(0.25, 0.5, 0.75, 1, 1.5, 2, 3, 5)) + 
  scale_color_manual(values = c("Significant (p < .05)" = "#1b9e77", "Not Significant" = "#7570b3")) +
  theme_minimal() +
  labs(
    title = "Multivariate Drivers of Ticket Resolution",
    subtitle = "Variables to the right (HR > 1) accelerate resolution. Variables to the left (HR < 1) delay it.",
    x = "Hazard Ratio (Log-Scaled Multiplier of Resolution Speed)",
    y = "Operational Variable",
    color = "Statistical Significance"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom",
    axis.text.y = element_text(face = "bold", size = 11)
  )
```

To interpret the multivariate forest plot, focus on the central dashed line at 1.00, which establishes the baseline resolution speed. The dots represent the Hazard Ratio—a direct speed multiplier for each operational variable. Variables resting to the right of the dashed line (HR > 1) accelerate resolution, while variables to the left (HR < 1) act as system bottlenecks that delay ticket closure. The horizontal bars attached to each point represent the statistical confidence interval; because none of these intervals cross the baseline, we can state with mathematical certainty that these variables genuinely impact workforce velocity rather than being random anomalies.

By explicitly filtering out the automated genear maximo integration to isolate true human labor, the model proves that resolving a municipal issue is heavily dependent on departmental capacity and intake friction. The Public Works department (PWDx) operates at nearly double the baseline speed, validating its structural design for rapid triage, whereas the Parks department (PARK) operates at a 35% speed deficit due to long-tail bottlenecks. Additionally, the data confirms that manual intake channels—such as constituent phone calls—introduce significant administrative friction, actively slowing down resolution pipelines by roughly 22% compared to more streamlined, digitally precise reporting.


# 9. System Stability: Queueing Theory and Backlog Velocity

While Survival Analysis evaluates the lifespan of individual tickets, it does not measure the aggregate thermodynamic health of a department. To diagnose systemic instability, we must shift from the ticket level to the macro system level using Queueing Theory.This approach relies strictly on the mathematical balance between the Arrival Rate ($\lambda$)—tickets opened—and the Service Rate ($\mu$)—tickets closed. By calculating the net weekly queue change, we can strip away subjective text and measure pure operational momentum. When $\lambda > \mu$, the system enters a state of structural instability (growing backlog), regardless of how fast routine tickets are closed.

```{r}
#| label: queue-velocity-analysis
#| echo: false
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Queueing Stability: Weekly net change in unresolved tickets (Arrivals minus Services)."

library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)

# 1. Calculate the Arrival Rate (lambda) and Service Rate (mu) per week
queue_dynamics <- sla_delta_data %>%
  filter(department %in% c("BTDT", "ISD", "INFO", "PARK", "PWDx")) %>%
  mutate(
    open_week = floor_date(open_dt, "week"),
    close_week = floor_date(closed_dt, "week")
  )

# Arrivals per week
arrivals <- queue_dynamics %>%
  group_by(department, week = open_week) %>%
  summarise(tickets_opened = n(), .groups = "drop")

# Services per week (ignoring NAs for open tickets)
services <- queue_dynamics %>%
  filter(!is.na(close_week)) %>%
  group_by(department, week = close_week) %>%
  summarise(tickets_closed = n(), .groups = "drop")

# 2. Merge and calculate Net Backlog Velocity
backlog_velocity <- arrivals %>%
  full_join(services, by = c("department", "week")) %>%
  # Replace NAs with 0 for weeks where no tickets were opened or closed
  replace_na(list(tickets_opened = 0, tickets_closed = 0)) %>%
  mutate(
    net_queue_change = tickets_opened - tickets_closed,
    # Create a flag for color coding: Red = Growing queue, Blue = Shrinking queue
    status = if_else(net_queue_change > 0, "Queue Growing (Instability)", "Queue Recovering (Stability)")
  ) %>%
  # Filter out extreme edge dates if necessary to keep the plot clean
  filter(week >= min(arrivals$week, na.rm = TRUE) & week <= max(arrivals$week, na.rm = TRUE))

# 3. Visualize the System Stability
ggplot(backlog_velocity, aes(x = week, y = net_queue_change, fill = status)) +
  geom_col(alpha = 0.85) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  facet_wrap(~ department, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("Queue Growing (Instability)" = "#e41a1c", 
                               "Queue Recovering (Stability)" = "#377eb8")) +
  theme_minimal() +
  labs(
    title = "Departmental Queue Stability (Backlog Velocity)",
    subtitle = "Tracking weekly thermodynamic balance: Tickets Opened vs. Tickets Closed.",
    x = "Week",
    y = "Net Change in Open Tickets",
    fill = "System Status"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

The Backlog Velocity chart maps the behavioral pulse of the city's workforce without relying on subjective administrative text. It reveals distinct operational realities across the monitored departments:

PARK (The Seasonal Unlock): For the first six months of the year, the Parks Department operates in an unbroken state of structural instability. The continuous red growth confirms they lack the baseline daily capacity to match inbound requests. The massive blue correction in July indicates a seasonal or fiscal-year budget unlock, allowing them to finally deploy the capital or seasonal workforce required to crush the six-month backlog.

BTDT (Administrative Batch Processing): Transportation displays a massive, singular correction of over 1,500 closed tickets in late summer. Because human workforces do not organically resolve 1,500 complex street-level issues in a single week, this is a clear artifact of database batch-closure. It represents a forced administrative reset to clear a runaway queue rather than natural operational velocity.

PWDx & INFO (Just-in-Time Reactivity): Public Works and the Mayor's Hotline constantly oscillate between small delays and immediate recoveries, closely hugging the baseline. This is the signature of highly reactive, well-managed workforces operating with zero structural slack. They absorb immediate shocks (such as the January weather surge for PWDx) and immediately deploy surge capacity the following week to reset to zero.

ISD (The Lagging Spring Surge): Inspectional Services demonstrates a steady accumulation of backlog through the spring months, directly aligning with the seasonal surge in construction and housing permits. Sustained recovery does not begin until mid-summer, indicating a persistent, multi-month lag between peak citizen demand and peak service delivery.

# 10. Spatial Equity (Geospatial Autocorrelation)

While overall departmental velocity provides a measure of system health, it does not address the distribution of that service. A critical mandate of municipal operations is equitable service delivery across all neighborhoods.

To mathematically test if operational failures are equitably distributed or geographically biased, we apply spatial autocorrelation using Global Moran’s I. By mapping the SLA breach rates (the percentage of tickets that miss their target resolution time) across Boston's spatial boundaries, we can identify statistically significant "cold spots"—clusters of neighborhoods that systematically receive delayed service compared to the citywide baseline.

## 10.1 Mapping SLA Breach Clusters

To mathematically test if operational failures are equitably distributed or geographically biased, we apply spatial autocorrelation using Global Moran’s I. The statistic compares the variance of SLA breach rates across neighboring areas against the overall citywide variance:$$I = \frac{N}{W} \frac{\sum_i \sum_j w_{ij}(x_i - \bar{x})(x_j - \bar{x})}{\sum_i (x_i - \bar{x})^2}$$An index near $0$ implies that service delays are completely random ("equitable failure"). A positive index approaching $1$ proves that SLA breaches are structurally clustered, creating "cold spots" of systematic neglect. To ensure we are measuring true human equity rather than software architecture, automated closures originating from the Maximo integration API have been strictly filtered out of this spatial model.

```{r}
#| label: spatial-equity-morans-i
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Geospatial distribution of SLA breach rates across Boston census tracts."

library(dplyr)
library(sf)
library(tigris)
library(spdep)
library(ggplot2)
library(stringr)

# 1. Conjure the Shapefile via API (Natively downloads Suffolk County Census Tracts)
options(tigris_use_cache = TRUE)
boston_tracts <- tracts(state = "MA", county = "Suffolk", cb = TRUE, year = 2022, progress_bar = FALSE) %>%
  st_transform(crs = 4326)

# 2. Prepare the 311 Spatial Points
spatial_311 <- sla_delta_data %>%
  # Keep only valid spatial coordinates
  filter(!is.na(latitude), !is.na(longitude)) %>%
  # Strip out the automated API artifacts to isolate human labor
  filter(!str_detect(if_else(is.na(source), "Unknown", as.character(source)), "Maximo")) %>%
  # Convert 'on_time' column into a binary breach flag. 
  # NOTE: Adjust "ONTIME" to precisely match the spelling/case in your dataset
  mutate(is_breach = if_else(on_time == "ONTIME", 0, 1)) %>% 
  # Convert flat data frame into an active spatial object
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# 3. Spatial Join: Assign tickets to Census Tracts and calculate breach rates
tract_breaches <- boston_tracts %>%
  st_join(spatial_311, left = FALSE) %>%
  group_by(GEOID) %>%
  summarise(
    total_tickets = n(),
    breach_rate = mean(is_breach, na.rm = TRUE)
  ) %>%
  # Filter out tracts with highly volatile low sample sizes (e.g., airport, harbor islands)
  filter(total_tickets > 50)

# 4. Calculate Global Moran's I for Spatial Autocorrelation
# Create the spatial weights matrix (defining which tracts touch each other)
neighbors <- poly2nb(tract_breaches, queen = TRUE)
weights <- nb2listw(neighbors, style = "W", zero.policy = TRUE)

# Execute the mathematical test
moran_result <- moran.test(tract_breaches$breach_rate, weights, zero.policy = TRUE)

# 5. Visualize the Inequity
ggplot(tract_breaches) +
  geom_sf(aes(fill = breach_rate), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "magma", labels = scales::percent_format()) +
  theme_void() +
  labs(
    title = "Spatial Equity of Municipal Services",
    subtitle = paste0("Global Moran's I: ", round(moran_result$estimate[1], 3), 
                      " (p-value: ", signif(moran_result$p.value, 3), ")"),
    fill = "SLA Breach Rate",
    caption = "Excludes automated Maximo Integration closures to isolate human workforce equity."
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "right"
  )
```

## 10.2 Localized Clusters of Neglect (LISA)

Having proven that municipal delays are not uniformly distributed, we apply Local Indicators of Spatial Association (LISA) to mathematically draw boundaries around the city's operational "hotspots" and "coldspots." By scaling the breach rates and calculating their spatial lags, we can isolate census tracts that suffer from statistically significant, systemic neglect.

Crucially, because automated closures originating from maximo integration have been removed from this spatial model, the resulting clusters represent genuine disparities in human workforce deployment and street-level service delivery.

```{r}
#| label: spatial-equity-lisa
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "LISA Cluster Map isolating statistically significant zones of systemic neglect (High-High) and exceptional service (Low-Low)."

# 1. Calculate Local Moran's I
local_m <- localmoran(tract_breaches$breach_rate, weights, zero.policy = TRUE)

# 2. Scale the variables to determine the quadrants (centered on the mean)
c_breach <- tract_breaches$breach_rate - mean(tract_breaches$breach_rate, na.rm = TRUE)
c_lag <- lag.listw(weights, c_breach, zero.policy = TRUE)

# 3. Categorize the tracts into LISA clusters
lisa_tracts <- tract_breaches %>%
  mutate(
    # Extract the p-value from the local moran matrix (Column 5)
    p_val = local_m[, 5], 
    
    # Define the raw spatial quadrants
    quadrant = case_when(
      c_breach > 0 & c_lag > 0 ~ "High-High (Systemic Neglect)",
      c_breach < 0 & c_lag < 0 ~ "Low-Low (Exceptional Service)",
      c_breach > 0 & c_lag < 0 ~ "High-Low (Spatial Outlier)",
      c_breach < 0 & c_lag > 0 ~ "Low-High (Spatial Outlier)",
      TRUE ~ "Not Significant"
    ),
    
    # Mask the quadrants by statistical significance (p < 0.05)
    lisa_cluster = if_else(p_val <= 0.05, quadrant, "Not Significant")
  ) %>%
  # Convert the cluster to a factor for ordered plotting
  mutate(lisa_cluster = factor(lisa_cluster, levels = c(
    "High-High (Systemic Neglect)", 
    "Low-Low (Exceptional Service)", 
    "High-Low (Spatial Outlier)", 
    "Low-High (Spatial Outlier)", 
    "Not Significant"
  )))

# 4. Map the LISA Clusters
# Define a stark, color-blind friendly palette to highlight the inequity
lisa_colors <- c(
  "High-High (Systemic Neglect)" = "#d73027",    # Red
  "Low-Low (Exceptional Service)" = "#4575b4",   # Blue
  "High-Low (Spatial Outlier)" = "#fdae61",      # Light Orange
  "Low-High (Spatial Outlier)" = "#abd9e9",      # Light Blue
  "Not Significant" = "#f0f0f0"                  # Light Grey
)

ggplot(lisa_tracts) +
  geom_sf(aes(fill = lisa_cluster), color = "darkgrey", size = 0.1) +
  scale_fill_manual(values = lisa_colors, drop = FALSE) +
  theme_void() +
  labs(
    title = "Zones of Systemic Neglect vs. Exceptional Service",
    subtitle = "Statistically significant spatial clusters of 311 SLA breaches (Local Moran's I, p < 0.05).",
    fill = "LISA Classification",
    caption = "Excludes automated genear maximo integration closures to isolate human workforce equity."
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(face = "bold")
  )
```